<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Project Status: Active – The project has reached a stable, usable
state and is being actively
developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![License: GPL
v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)

# s-jSDM (Mojo fork) - Joint Species Distribution Modeling accelerated on Apple Silicon CPUs

## What this repository is

This is a fork of [TheoreticalEcology/s-jSDM](https://github.com/TheoreticalEcology/s-jSDM)
that replaces the inner Monte Carlo likelihood computation of sjSDM with a
hand-written **Mojo/MAX kernel**, giving roughly **1.3--3x faster model fits
on Apple Silicon (M-series) CPUs** while producing statistically equivalent
results.

**This repository is only for users who want the Mojo-accelerated version of
sjSDM on an Apple Silicon Mac.** It currently requires:

- An Apple Silicon Mac (arm64 macOS)
- A clone of this repository (the Mojo worker binary is built locally)

If you are not in that audience -- Intel Macs, Windows, or Linux -- please
use the **original package instead**:

```r
remotes::install_github("TheoreticalEcology/s-jSDM/sjSDM")
library(sjSDM)
install_sjSDM()
```

**On Apple Silicon but don't want Mojo?** Stay here anyway. Installing
sjSDM from the upstream repository is currently unreliable on Apple
Silicon, whereas this fork contains the compatibility fixes (released as
[v0.1.0](https://github.com/dougwyu/s-jSDM/releases/tag/v0.1.0)). You can
use plain PyTorch-mode sjSDM from this fork without any of the pixi/Mojo
setup:

```r
remotes::install_github("dougwyu/s-jSDM/sjSDM", ref = "v0.1.0")
library(sjSDM)
install_sjSDM()
```

Everything about fitting and interpreting models is identical between all
of these; only the compute backend differs.

## What sjSDM does

sjSDM is an R package for estimating joint species distribution models
(jSDMs): GLMMs that model a many-species response to the environment,
space, and a covariance term capturing conditional correlations between
species. Unlike latent-variable approximations, sjSDM fits a full
(low-rank) covariance matrix in the likelihood, numerically approximated
via simulations. The method is described in Pichler & Hartig (2021),
*A new joint species distribution model for faster and more accurate
inference of species associations from big community data*,
https://www.doi.org/10.1111/2041-210X.13687.

The core computations run in Python/PyTorch wrapped into R; this fork adds
a native Mojo kernel alongside PyTorch for the hot Monte Carlo loop.
Citation info: `citation("sjSDM")`.

## Installation (Mojo backend)

One-time setup -- install [pixi](https://pixi.sh), create the pinned Python
environment, build the Mojo worker, and create a thin venv wrapper that
lets recent versions of `reticulate` use the pixi environment:

```bash
git clone https://github.com/dougwyu/s-jSDM.git
cd s-jSDM/work/mojo-backend
pixi install
pixi run mojo build -O 3 mojo/mc_logit_grad_server.mojo -o mc_grad_server_bin
.pixi/envs/default/bin/python -m venv reticulate-venv --system-site-packages
```

The `-O 3` flag matters: unoptimized builds are ~50x slower. Binaries embed
absolute paths, so redo these commands after moving or re-cloning the
repository. The final command is needed because `reticulate` >= 1.45
refuses conda-style environments without a `conda` binary and silently
falls back to an environment without PyTorch; the wrapper venv reuses the
pixi environment's packages via `--system-site-packages`.

In R, load the package from the checkout (this is the supported way to use
the fork -- see the note below on why not `library(sjSDM)`):

```{r setup-load, eval=FALSE}
Sys.setenv(RETICULATE_PYTHON = here::here(
  "work/mojo-backend/reticulate-venv/bin/python"
))
devtools::load_all(here::here("sjSDM"), quiet = TRUE)
is_torch_available()
```

> **Why not `library(sjSDM)`?** If you install this fork as a regular
> package, the installed copy looks for the Mojo worker *inside your R
> library*, won't find it, and silently falls back to PyTorch -- no error,
> just no acceleration. Just use `devtools::load_all()` on the checkout as
> shown above.

That's it -- the Mojo backend is used **automatically** whenever the worker
binary exists and the fit runs on CPU float32 logit/probit. Control it with
one environment variable:

```{r mojo-switch, eval=FALSE}
Sys.setenv(SJSDM_MOJO_BACKEND = "auto")  # default: Mojo when available
Sys.setenv(SJSDM_MOJO_BACKEND = "0")     # force pure PyTorch
Sys.setenv(SJSDM_MOJO_BACKEND = "1")     # force Mojo (error if missing)
```

Standard errors / Hessians always use PyTorch regardless of the setting.

For a complete walkthrough of every feature -- fitting, species
associations, prediction, variation partitioning, cross-validation, spatial
models, plots, benchmarking torch vs. Mojo, and troubleshooting -- see
[Code/sjSDM_mojo_tutorial.Rmd](Code/sjSDM_mojo_tutorial.Rmd).

## Basic workflow

```{r workflow, eval=FALSE}
# after the setup chunk above (load_all + RETICULATE_PYTHON):
set.seed(42)
community <- simulate_SDM(sites = 100, species = 10, env = 3)
Env <- community$env_weights
Occ <- community$response

model <- sjSDM(Y = Occ,
               env = linear(data = Env, formula = ~X1+X2+X3),
               se = TRUE, family = binomial("probit"),
               sampling = 100L, verbose = FALSE)

summary(model)
plot(model)          # niche estimates with error bars
image(getCor(model)) # species association matrix
anova(model)         # variation partitioning
```

## Validating statistical accuracy

The Mojo kernel uses different float32 arithmetic than PyTorch, so fitted
models are statistically equivalent rather than identical. To check that
equivalence quantitatively, the repository ships a validation script that
fits both backends to synthetic communities with known ground truth and
compares how well each recovers it, across three targets: environmental
coefficients, the species association matrix, and per-species loadings on
an autocorrelated spatial field (dispersal limitation):

```{r accuracy-test, eval=FALSE}
Rscript Code/test_statistical_accuracy.R          # ~10 min, 3 replicates
```

It prints a table of recovery metrics per backend (correlation with truth,
attenuation slope, RMSE) and exits non-zero if the backends ever disagree
beyond tolerance. Reference result from the current build:

| metric | torch | mojo | worst gap |
|---|---|---|---|
| env coefficient correlation | 0.964 | 0.966 | 0.008 |
| association matrix correlation | 0.807 | 0.812 | 0.018 |
| spatial loading correlation | 0.988 | 0.986 | 0.010 |

## Notes and caveats

- Cross-backend results are statistically equivalent, not bit-identical:
  float32 reduction-order differences perturb trajectories slightly.
- Keep predictors standardized (`scale(env)`); fits can diverge on raw-scale
  covariates under either backend.
- The Mojo backend is CPU-only. GPU users can still set
  `options(sjSDM.device = "mps")`, which routes fits through PyTorch MPS.
- Releases: [v0.2.0 "Mojo CPU acceleration"](https://github.com/dougwyu/s-jSDM/releases/tag/v0.2.0)
  is the first tagged release of the fork.

## License

GPL-3, matching the upstream project. The Mojo kernel sources live under
`work/mojo-backend/mojo/`.

## Acknowledgments

This port was carried out with the assistance of **Ox Alpha**, an AI coding
and data-analysis assistant developed by an undisclosed organization.
Ox Alpha implemented and validated the Mojo/MAX kernels, the persistent
worker bridge, the SIMD optimizations, and much of the benchmarking and
parity-testing infrastructure described in this repository, working in
collaboration with Douglas Yu. All AI-generated changes were reviewed,
tested, and committed by the repository maintainer.
