---
title: "sjSDM from R: joint species distribution models and Mojo CPU setup"
author: "sjSDM developers"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
# Set run_live to TRUE to knit this document with real fits, tables, and
# figures. It is FALSE by default because the fits need the reticulate/pixi
# environment described in the Installation section, and a few chunks
# (cross-validation, standard errors) are slow.
run_live <- FALSE

knitr::opts_chunk$set(
  eval       = run_live,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.2,
  dpi        = 110
)
```

This tutorial has three parts:

1.  Understand what a joint species distribution model is, what it estimates, and what its output can and cannot support.
2.  Install and select either the standard PyTorch backend or the optional Mojo CPU backend.
3.  Work through the main sjSDM workflow: simulation, fitting, checking that the model recovers known truth, species associations, assembly-process tests, prediction, $R^2$, variation partitioning, cross-validation, spatial models, standard errors, and plotting.

Code chunks do not run while this document is rendered, unless you set `run_live <- TRUE` in the setup chunk. Run them interactively, in order. The basic fit takes a second or two; cross-validation, ANOVA refits, and standard errors take longer.

If you already know jSDM theory and only want the software, skip to the Installation section. If you have never fitted one, read the next section first, because most of the mistakes people make with these models are conceptual rather than computational.

# Joint species distribution models: the concepts

## From one species at a time to all species at once

A *species distribution model* (SDM) relates the occurrence of one species to environmental predictors. Fit one per species and you have a *stacked* SDM: a collection of independent models. Stacking is easy, and for pure prediction of one species at a time it is often fine. What it cannot do is say anything about how species co-occur beyond what the environment already explains, because it assumes, by construction, that the species are independent once you condition on the covariates.

A *joint* SDM (jSDM) fits all species simultaneously and adds one thing: a **residual correlation matrix** across species. After removing the environmental signal, do species still tend to occur together, or to exclude each other? That residual structure is the extra parameter, and it is the reason to use a jSDM at all.

Two practical consequences follow immediately.

- **Species with few records borrow strength.** A rare species with 12 occurrences gets a badly determined coefficient on its own. In a joint model its latent factor loadings are estimated using the co-occurrence pattern it shares with better-sampled species, which stabilises the fit.
- **Predictions become conditional.** With a joint model you can ask "given that species A is present here, how likely is species B?" A stacked model cannot answer that question, because it has already assumed the answer is "no different from usual".

## A worked example: hidden structure in a ratings matrix

This subsection needs no ecology and no linear algebra beyond multiplying two matrices. It is here because the model in the next subsection is much easier to read once you have seen the same idea solve a problem you already understand.

### The problem

A streaming service keeps one row per viewer and one column per film. A cell holds 1 if that viewer liked the film, 0 if they did not, and `NA` if they never watched it. Fifteen viewers, twenty-five films, and here is a corner of it.

```{r films-data, echo=FALSE}
set.seed(3)
viewers <- c("Ana","Ben","Cara","Dan","Eve","Femi","Gita","Hugo",
             "Iris","Jonas","Kemi","Lars","Mira","Noor","Omar")
films <- c("Die Hard","Mad Max","John Wick","Terminator","Speed",
           "Top Gun","Predator","Face/Off","Point Break",
           "Notting Hill","Love Actually","Harry Met Sally","Pretty Woman",
           "Sleepless","Bridget Jones","Four Weddings","The Notebook",
           "Stalker","Persona","Seventh Seal","Mulholland Dr",
           "In Mood for Love","Tokyo Story","Wild Strawberries","Blue Velvet")
genre <- rep(c("action","romance","arthouse"), times = c(9, 8, 8))

V_true <- matrix(0, 25, 3, dimnames = list(films, c("action","romance","arthouse")))
for (k in 1:3) V_true[genre == colnames(V_true)[k], k] <- 1
V_true <- V_true + matrix(rnorm(75, 0, 0.12), 25, 3)
U_true <- matrix(rnorm(45, 0, 1), 15, 3, dimnames = list(viewers, colnames(V_true)))

liked   <- matrix(rbinom(375, 1, plogis(6 * U_true %*% t(V_true))), 15, 25,
                  dimnames = list(viewers, films))
watched <- matrix(runif(375) < 0.55, 15, 25, dimnames = dimnames(liked))
ratings <- liked
ratings[!watched] <- NA

ratings[1:8, c("Die Hard","Top Gun","Speed","Notting Hill",
               "Pretty Woman","Sleepless","Stalker","Persona")]
```

Only 207 of the 375 cells are filled in. The job is to fill an empty one: find a film a viewer has not seen that they would like. Here is what Ana has left.

```{r films-ana-unseen, echo=FALSE}
films[is.na(ratings["Ana", ])]
```

### Why this is harder than it looks

The obvious idea is that people like films resembling ones they already liked. That immediately raises a question: resembling how? Nobody has labelled these films, and even a genre label might not be the thing that predicts taste.

So take the similarity from the ratings themselves. Two films are alike if the same people liked both. That works, but the thing you actually want runs the other way: find viewers like Ana, and recommend what *they* liked that Ana has not seen. Two viewers are alike if they liked the same films.

Look at those two sentences together. Film similarity is defined through viewers, and viewer similarity is defined through films. Neither can be worked out first, because each needs the other. That circle is the real problem.

Factoring the matrix breaks it, by solving for viewers and films at the same time.

### Factoring the matrix

Suppose every film had a position on a small number of hidden dimensions, and every viewer had a taste score on those same dimensions. Then a rating would just be the two lined up: multiply the scores pair by pair and add. If a film scores high on a dimension the viewer also scores high on, the rating goes up.

In matrix terms, a 15 by 25 table of ratings would be reproduced by a 15 by 3 table of viewer tastes multiplied by a 3 by 25 table of film positions. That is the factorisation, and 3 is a choice you make.

You know neither table. But notice: if you *knew* the film positions, finding one viewer's tastes would be an ordinary regression of their observed ratings on the positions of the films they rated. And if you knew every viewer's tastes, each film's position would be the same kind of regression over the viewers who rated it. So guess one, solve for the other, and go back and forth until it settles. That is the entire algorithm.

```{r films-factorise}
factorise <- function(R, K, lambda = 0.3, iters = 60) {
  obs <- !is.na(R)
  set.seed(1)
  U <- matrix(rnorm(nrow(R) * K, 0, 0.1), nrow(R), K)   # viewer tastes, rough guess
  V <- matrix(rnorm(ncol(R) * K, 0, 0.1), ncol(R), K)   # film positions, rough guess
  I <- diag(K)

  for (it in seq_len(iters)) {
    for (i in seq_len(nrow(R))) {        # each viewer, from the films they rated
      j <- which(obs[i, ])
      Vj <- V[j, , drop = FALSE]
      U[i, ] <- solve(crossprod(Vj) + lambda * I, crossprod(Vj, R[i, j]))
    }
    for (j in seq_len(ncol(R))) {        # each film, from the viewers who rated it
      i <- which(obs[, j])
      Ui <- U[i, , drop = FALSE]
      V[j, ] <- solve(crossprod(Ui) + lambda * I, crossprod(Ui, R[i, j]))
    }
  }
  dimnames(U) <- list(rownames(R), paste0("d", 1:K))
  dimnames(V) <- list(colnames(R), paste0("d", 1:K))
  list(U = U, V = V, fitted = U %*% t(V))
}

mf <- factorise(ratings, K = 3)
score <- mf$fitted
dimnames(score) <- dimnames(ratings)
```

Only observed cells enter either regression, so the `NA` cells are skipped rather than guessed at. The `lambda * I` term keeps a viewer with nine ratings from getting extreme coefficients. And `score` is now filled in everywhere, including the cells that were empty, because once you have both tables every cell is just a multiplication.

So, for Ana:

```{r films-ana-top}
unseen <- which(is.na(ratings["Ana", ]))
round(sort(score["Ana", unseen], decreasing = TRUE)[1:3], 2)
```

Does that work in general? Take each viewer's three highest-scoring unseen films and check whether they would in fact have liked them.

```{r films-check}
top3 <- function(v) {
  unseen <- which(is.na(ratings[v, ]))
  head(unseen[order(score[v, unseen], decreasing = TRUE)], 3)
}
round(c(`top-3 hit rate` = mean(sapply(viewers, function(v) mean(liked[v, top3(v)]))),
        `base rate`      = mean(liked[is.na(ratings)])), 3)
```

84% of the recommended films are liked, against 49% for unseen films picked at random.

### What the hidden dimensions turn out to be

The interesting part is not the recommendations. It is that the three dimensions mean something. Look at which films score highest on each.

```{r films-raw-axes}
for (k in 1:3) cat(sprintf("d%d: %s\n", k,
    paste(head(films[order(mf$V[, k], decreasing = TRUE)], 4), collapse = "; ")))
```

That is disappointing. The first dimension puts *Bridget Jones* next to *Terminator*, which is not a genre anybody would name.

The reason is worth knowing. The factorisation pins down the *product* of the two tables, not the tables themselves. You can rotate the dimensions, apply the same rotation to viewers and films alike, and get an identical fit. The algorithm returned one orientation out of infinitely many, and it had no reason to pick an interpretable one.

Since the choice is free, choose it. `varimax()` rotates the axes to make each one load heavily on a few films and near zero on the rest, which is the usual way to get interpretable factors.

```{r films-rotate}
name_axes <- function(f) {
  rt <- varimax(f$V)$rotmat
  V <- f$V %*% rt
  U <- f$U %*% rt
  s <- sign(colSums(V)); s[s == 0] <- 1       # point each axis the positive way
  V <- sweep(V, 2, s, "*"); U <- sweep(U, 2, s, "*")
  dimnames(V) <- dimnames(f$V); dimnames(U) <- dimnames(f$U)
  list(U = U, V = V)
}

rot <- name_axes(mf)
max(abs(rot$U %*% t(rot$V) - mf$fitted))      # the fit itself has not changed

for (k in 1:3) cat(sprintf("axis %d: %s\n", k,
    paste(head(films[order(rot$V[, k], decreasing = TRUE)], 5), collapse = "; ")))
```

Now read them off. Axis 1 is romance, axis 2 is arthouse, axis 3 is action. Nobody supplied a genre label, an actor, or a plot summary. The only input was a half-empty grid of ones and zeros.

The same three numbers describe the viewers, and now they can be read as taste.

```{r films-viewers}
tastes <- rot$U
colnames(tastes) <- c("romance", "arthouse", "action")
round(tastes[1:8, ], 2)
```

Dan is a romance viewer and nothing else. Femi is there for action. Ana is high on arthouse and action but flat on romance, which is exactly why the recommender offered her *Blue Velvet*, *Persona*, and *Face/Off* rather than *The Notebook*.

One caution before moving on. It is the *set* of axes, or equivalently their cross-product, that the data pin down. Any particular axis is a convention you imposed by choosing varimax, so "axis 1 is romance" is a reading of this rotation, not a fact the fit established.

### Where the matrix came from

The ratings were simulated, so we can check the answer. Here is the code that was hidden at the top.

```{r films-source, ref.label="films-data", eval=FALSE}
```

Each film was assigned to one of three genres, given a position near 1 on that genre's dimension and near 0 on the others, and each viewer was given three random taste scores. The probability of liking came from multiplying the two, and a coin flip turned it into 0 or 1. The recovered axes match the genres that went in, which is the check that the method found real structure rather than an artefact.

### The same matrix in ecology

Now change the nouns and nothing else. Rows are survey sites, columns are species, and a cell is 1 if the species was recorded there and 0 if not.

```{r birds-data, echo=FALSE}
set.seed(11)
sites <- c("Ashdown","Beckley","Cranmere","Dunwich","Elmsett","Foxholes",
           "Granborough","Hartop","Ickworth","Jurby","Kelling","Langmere",
           "Marden","Netherby","Oakley")
species <- c("Great Tit","Blue Tit","Nuthatch","Treecreeper","Blackcap","Chiffchaff",
             "Jay","Woodpecker","Marsh Tit",
             "Reed Warbler","Sedge Warbler","Coot","Moorhen","Little Grebe",
             "Water Rail","Snipe","Reed Bunting",
             "Skylark","Yellowhammer","Linnet","Corn Bunting","Lapwing",
             "Grey Partridge","Whitethroat","Meadow Pipit")
habitat <- rep(c("woodland","wetland","farmland"), times = c(9, 8, 8))

L_true <- matrix(0, 25, 3, dimnames = list(species, c("woodland","wetland","farmland")))
for (k in 1:3) L_true[habitat == colnames(L_true)[k], k] <- 1
L_true <- L_true + matrix(rnorm(75, 0, 0.12), 25, 3)
E_true <- matrix(rnorm(45, 0, 1), 15, 3, dimnames = list(sites, colnames(L_true)))

Y <- matrix(rbinom(375, 1, plogis(3 * E_true %*% t(L_true))), 15, 25,
            dimnames = list(sites, species))

Y[1:6, c("Blue Tit","Nuthatch","Jay","Coot","Snipe","Skylark","Lapwing","Linnet")]
```

This matrix has no `NA` cells, because every site was fully surveyed. That is the usual ecological case, and it means we are no longer filling gaps. We run exactly the same function anyway, for a different reason.

```{r birds-factorise}
bf  <- factorise(Y, K = 3)
brot <- name_axes(bf)

for (k in 1:3) cat(sprintf("axis %d: %s\n", k,
    paste(head(species[order(brot$V[, k], decreasing = TRUE)], 5), collapse = "; ")))
```

Axis 1 is a wetland axis, axis 2 is farmland, axis 3 is woodland. Nobody recorded habitat. There is no habitat column anywhere in `Y`, which contains only which bird was seen where. The three axes were reconstructed from co-occurrence alone.

The site scores say what each site is.

```{r birds-sites}
habitats <- brot$U
colnames(habitats) <- c("wetland", "farmland", "woodland")
round(habitats, 2)
```

Jurby is woodland and little else. Hartop, Dunwich, and Beckley are farmland. Langmere and Ickworth are wet. Granborough and Ashdown are mixed, scoring on everything.

### What this means for a jSDM

That last result is the whole motivation for the model in the next subsection, so it is worth stating plainly.

A latent dimension is an unmeasured gradient along which sites differ and species respond. Here it happened to be habitat. In a real dataset it could be soil pH, canopy openness, grazing history, or a survey artefact. You do not observe it, you do not name it in advance, and the method finds it anyway, because species that respond to the same hidden gradient co-occur.

Three consequences follow directly.

- **Species that load on the same axis co-occur.** Turn the species loadings into a species-by-species matrix and you have exactly the **species association matrix** that sjSDM reports through `getCov()` and `getCor()`. It is the ecological version of "which films go together".
- **A strong association is not evidence of interaction.** Nothing in this bird example involves species interacting. They share a habitat preference. That is the warning from the previous subsection, now visible as arithmetic: a missing covariate and a biotic interaction produce the same pattern.
- **Measuring the gradient makes it go away.** Had we recorded habitat and put it in as a predictor, that structure would move out of the latent part and into the measured part. This is why a jSDM has both: $\mathbf{X}\boldsymbol{\beta}$ for what you measured, and a low-rank latent term for what you did not. Adding a covariate and watching an association shrink is the model working, not a contradiction.

Here is the ecological data-generating code, for the same check as before.

```{r birds-source, ref.label="birds-data", eval=FALSE}
```

### How sjSDM stores it

Two differences from the tables above will otherwise trip you up when you look at a fitted object.

**The film-position table is kept untransposed.** Above, `rot$V` is films by dimensions, 25 by 3, and the product needs `t(V)`. sjSDM's `sigma` is likewise species by `df`, so with `df = 5` and 10 species it is 10 by 5. That is why `getCov()` is computed as `sigma %*% t(sigma)`: it is the species-by-species cross-product, the same object as the film similarity above.

**The site scores are never stored.** This is the real departure. A recommender must estimate `U`, because recommending to Ana requires Ana's taste vector. sjSDM does not: site scores are redrawn from a standard normal every time the likelihood is evaluated and averaged over, so a fitted model keeps `sigma` and nothing else. The two lines that do it in the PyTorch path, quoted here rather than run, are

``` python
# Python, from sjSDM/inst/python/sjSDM_py/model_sjSDM.py
noise = torch.randn(size = [sampling, batch_size, df])      # site scores, redrawn
torch.einsum("ijk, lk -> ijl", [noise, sigma])              # noise %*% t(sigma)
```

where `noise` is samples by sites by `df` and `sigma` is species by `df`. The einsum multiplies and sums over the shared `df` index, which is `U %*% t(V)` done for `sampling` draws at once.

Two consequences, both cashed out in the next subsection. Because the scores are standard normal rather than free parameters, the only remaining freedom is a rotation, exactly as with the film axes, so the association matrix is identified even though the loadings are not. And because they are averaged over rather than estimated, the likelihood is an integral, which is where `sampling` and the Monte Carlo machinery come from.

## The model sjSDM fits

For site $i$ and species $j$, sjSDM models the observed occurrence $y_{ij}$ as

$$y_{ij} \sim \text{Bernoulli}(p_{ij}), \qquad g(p_{ij}) = \beta_{0j} + \mathbf{x}_i^\top \boldsymbol{\beta}_j + e_{ij}$$

where $g$ is the link (logit or probit), $\mathbf{x}_i$ are the environmental predictors at site $i$, and $\boldsymbol{\beta}_j$ are that species' environmental responses. The last term is the joint part: the site-level residual vector $\mathbf{e}_i = (e_{i1}, \dots, e_{iJ})$ is drawn once per site from a multivariate normal shared by all species,

$$\mathbf{e}_i \sim \mathcal{N}_J(\mathbf{0}, \boldsymbol{\Sigma}), \qquad \boldsymbol{\Sigma} = \boldsymbol{\Lambda}\boldsymbol{\Lambda}^\top .$$

$\boldsymbol{\Lambda}$ is a $J \times d$ matrix of factor loadings, where $d$ is the rank you choose with `bioticStruct(df = d)`. Writing $\mathbf{e}_i = \boldsymbol{\Lambda}\mathbf{z}_i$ with $\mathbf{z}_i \sim \mathcal{N}_d(\mathbf{0}, \mathbf{I})$ gives the equivalent *latent variable* form used in the HMSC and boral literature: $d$ unmeasured site-level gradients, each species loading onto them with its own weights.

Three things to take from this.

- **The low rank is the whole trick.** A free $J \times J$ covariance has $J(J+1)/2$ parameters, which for 200 species is over 20,000 and is hopeless to estimate. The factorisation reduces it to $J \times d$. With `df = 5` and 200 species that is 1,000 parameters instead. You choose `df`; it is a modelling decision, not a nuisance.
- **The latent scale is not fixed.** sjSDM does not constrain $\text{diag}(\boldsymbol{\Sigma})$ to 1, so the residual variance is free and trades off against the size of $\boldsymbol{\beta}$. This is why `getCov()` returns diagonal entries well above 1, and why you should compare estimated **correlations**, not covariances, against a known truth. The recovery section below does exactly that.
- **The likelihood has no closed form.** Under the low-rank representation, the site likelihood integrates over $\mathbf{z}_i$, a $d$-dimensional standard Gaussian latent variable with no analytic solution for either link. Each draw is mapped through $\boldsymbol{\Lambda}\mathbf{z}_i$ into a rank-$d$ residual in the $J$-dimensional species space. sjSDM draws `sampling` values of $\mathbf{z}_i$ per site per minibatch and averages the resulting likelihoods. That inner loop is where nearly all the compute goes, and it is exactly the loop the optional Mojo backend replaces.

## What a residual correlation does and does not mean

This is the single most important idea in the tutorial, and the one most often got wrong in published applications.

A positive residual correlation between two species means: *after accounting for the environmental predictors you supplied*, these species co-occur more than independence would predict. It does **not** mean they interact.

At least four things produce the same signal.

- A genuine biotic interaction: facilitation, shared mutualists, predator and prey.
- A **missing environmental covariate**. If both species track soil pH and you did not measure soil pH, their shared response to it lands in the residual correlation. This is by far the most common cause in real data.
- **Shared spatial structure**, from dispersal limitation or historical biogeography, that no measured variable captures.
- **Sampling artefacts**: a shared detection bias, a shared observer, a shared season.

The practical consequence is that residual correlations are a description of what your covariates failed to explain, not a measurement of ecological interaction. Adding a covariate can make an association vanish, and that is the model working correctly, not a contradiction. The honest way to report one is "residual association, conditional on this set of predictors", and the honest way to strengthen an interaction claim is to show that the association survives the addition of plausible confounders, including space.

The dispersal-limitation and environmental-filtering sections below are built to make this concrete: they show the same association machinery picking up structure that is not an interaction at all.

## Four assembly processes, and which ones a jSDM can separate

Community ecology conventionally names four processes that determine which species occur where: environmental filtering, biotic interactions, dispersal, and stochasticity. A jSDM addresses them unevenly, and it helps to be explicit about which is which.

- **Environmental filtering** is the $\mathbf{x}_i^\top \boldsymbol{\beta}_j$ term. This is what the model estimates best, and variation partitioning reports it as the "Abiotic" component.
- **Dispersal** is partly addressable, but only if you give the model spatial predictors. `generateSpatialEV()` builds Moran eigenvector maps from coordinates, and `sjSDM(spatial = ...)` fits them as a separate component so the partition reports "Spatial" alongside "Abiotic". Without spatial predictors, spatial structure has nowhere to go except the residual correlation.
- **Biotic interactions** are what people hope $\boldsymbol{\Sigma}$ measures. It measures residual association, which is an upper bound on the interaction signal and is confounded with everything in the previous subsection.
- **Stochasticity** is the unexplained remainder, and no partitioning separates it from model misspecification.

So a fair summary of what the machinery delivers: a good estimate of environmental response, a decomposition of explained deviance into environment, space, and residual association, and a residual association matrix that requires careful interpretation.

## Practical guidance before you fit anything

- **Standardize your predictors.** sjSDM fits can diverge on raw-scale covariates, for example temperatures near 6000. Use `scale()`. This affects both backends and is the most common cause of a fit that goes nowhere.
- **Watch prevalence.** Species present at fewer than roughly 10 sites contribute very little and can destabilise the association estimates. Dropping them is defensible; report that you did.
- **Sites matter more than species.** The residual correlation is estimated across sites, so 500 sites and 30 species supports far better association inference than 30 sites and 500 species, even though the second matrix is larger.
- **Choose `df` deliberately.** It is the number of latent gradients. A small value (2 to 5) is a strong regulariser and is a reasonable default; increasing it lets the model represent more association structure at the cost of variance. Tune it, or `lambda`, by cross-validation rather than by eye.
- **Expect shrinkage.** Estimated correlations are systematically smaller in magnitude than the truth, for reasons the recovery section quantifies. Compare patterns and orderings across datasets rather than treating a single coefficient as calibrated.
- **Set a seed and report it.** The likelihood is stochastic. Two runs of the same model on the same data will not be identical.

## Where each idea lives in the sjSDM API

- The response matrix $\mathbf{Y}$: the `Y` argument, sites by species, 0/1.
- Environmental responses $\boldsymbol{\beta}$: `env = linear(data = X, formula = ~ ...)`, or `DNN()` for a non-linear predictor.
- Spatial component: `spatial = linear(data = SP, formula = ~ 0 + .)` with `SP` from `generateSpatialEV()`.
- Association structure $\boldsymbol{\Lambda}$: `biotic = bioticStruct(df = d, lambda = ..., on_diag = ...)`.
- Link $g$: `family = binomial("logit")` or `binomial("probit")`.
- Monte Carlo effort: `sampling`; optimisation effort: `iter` and `step_size`.
- Fitted associations: `getCov()` and `getCor()`. Fitted coefficients: `coef()`. Partition: `anova()`.

The optional **Mojo backend** replaces the inner Monte Carlo likelihood computation with a hand-written kernel compiled to native Apple Silicon code. Predictor layers, optimizers, prediction, and Hessian-based standard errors remain in PyTorch. It changes speed, not the model.

# Installation

## Who this tutorial is for

This tutorial covers the **Mojo-accelerated CPU build** of sjSDM, which currently requires **Apple Silicon (macOS, arm64)** and a full clone of the development repository at `dougwyu/s-jSDM`.

> **Not on Apple Silicon?** You don't need this tutorial. Install the original, pure-PyTorch sjSDM instead -- it is fully supported on macOS (Intel), Windows, and Linux:
>
> ```{r fallback, eval=FALSE}
> install.packages("remotes")
> remotes::install_github("TheoreticalEcology/s-jSDM/sjSDM")
> library(sjSDM)
> install_sjSDM(version = "cpu")
> install_diagnostic()   # use this detailed report if something is off
> ```
>
> **On Apple Silicon but don't want Mojo?** Stay here anyway. Installing from the upstream repository is currently unreliable on Apple Silicon, whereas this fork contains the compatibility fixes (released as [v0.1.0](https://github.com/dougwyu/s-jSDM/releases/tag/v0.1.0)). Plain PyTorch-mode sjSDM works without any of the pixi/Mojo setup below:
>
> ```{r fallback-apple, eval=FALSE}
> install.packages("remotes")
> remotes::install_github("dougwyu/s-jSDM/sjSDM", ref = "v0.1.0")
> library(sjSDM)
> install_sjSDM(version = "cpu")
> install_diagnostic()   # use this detailed report if something is off
> ```
>
> Everything from the simulation section onward works identically with either install; just skip the Mojo setup and switching sections.

## One install path: clone + pixi

**If everything is already installed, jump to Step 4.**

The Mojo backend lives alongside a project-local Python environment defined with [pixi](https://pixi.sh), a single-binary package manager. PyTorch 2.5.1, Mojo 1.0.0, and MAX 26.5.0 are the versions validated for v0.2.1. The ranges in `pixi.toml` are not an immutable lock, so commit a tested `pixi.lock` before relying on byte-for-byte reproducible installs.

**Step 1 -- clone the repository:**

```{bash clone, eval=FALSE}
git clone https://github.com/dougwyu/s-jSDM.git
cd s-jSDM
```

**Step 2 -- install pixi** (one-time; see <https://pixi.sh> for alternatives):

```{bash pixi, eval=FALSE}
# macOS/Linux:
curl -fsSL https://pixi.sh/install.sh | sh
```

**Step 3 -- create the Python environment, build the Mojo worker, and create the reticulate wrapper.** From the repository root:

```{bash build, eval=FALSE}
cd work/mojo-backend
pixi install
pixi run mojo build -O 3 mojo/mc_logit_grad_server.mojo -o mc_grad_server_bin
.pixi/envs/default/bin/python -m venv reticulate-venv --system-site-packages
```

The `-O 3` flag matters: unoptimized Mojo builds are roughly 50x slower. The thin `reticulate-venv` wrapper is also required: recent reticulate versions do not treat a pixi prefix without a `conda` executable as a usable Conda environment. When this step finishes, both of these paths should exist:

```{bash verify-build, eval=FALSE}
test -x mc_grad_server_bin
test -x reticulate-venv/bin/python
```

**Jump here if everything is already installed**

**Step 4 -- install the R-side development tools, select a backend, and load the package.** Run the package installation once:

```{r r-prerequisites, eval=FALSE}
install.packages(c("devtools", "here", "mvtnorm", "knitr"))
```

Then start each R session from the repository and run:

```{r reticulate}
Sys.setenv(RETICULATE_PYTHON = here::here(
  "work/mojo-backend/reticulate-venv/bin/python"
))
Sys.setenv(SJSDM_MOJO_BACKEND = "auto")
stopifnot(file.exists(Sys.getenv("RETICULATE_PYTHON")))

# load_all() runs the REPO code -- including the Python backend in
# sjSDM/inst/python AND its ability to find the Mojo worker binary:
devtools::load_all(here::here("sjSDM"), quiet = TRUE)

is_torch_available()
reticulate::py_config() # use this diagnostic with the pixi setup
```

> **Why not `library(sjSDM)`?** If you install this fork as a regular package, the installed copy looks for the Mojo worker *inside your R library*, won't find it, and silently falls back to PyTorch -- no error, just no acceleration. Just use `devtools::load_all()` on the checkout as shown above.

> **Key idea for R users:** R talks to Python through `reticulate`, which sjSDM handles internally. `RETICULATE_PYTHON` selects the Python environment. The Mojo binary is a separate native worker that sjSDM finds automatically; you do not call it directly.

# Using the Mojo CPU backend

## What it is

The Mojo backend accelerates the Monte Carlo logit/probit likelihood and its analytic gradients on **CPU**. It consists of the two pieces you built during installation:

- **Persistent worker**, `work/mojo-backend/mc_grad_server_bin`: a long-lived process that computes loss and gradients for each training batch.
- **R-side bridge**, `sjSDM/inst/python/sjSDM_py/mojo_bridge.py`: feeds tensors and seeds to the worker through a pipe.

You do **not** call either directly. Backend selection is read from Python's environment when a model is built.

## Turning it on / off

Set the backend **before** `devtools::load_all()` starts Python:

```{r mojo-switch, eval=FALSE}
# Pick ONE of these; they are alternatives, not a sequence.
Sys.setenv(SJSDM_MOJO_BACKEND = "auto") # default: Mojo when eligible
Sys.setenv(SJSDM_MOJO_BACKEND = "1")    # force Mojo; error if unavailable
Sys.setenv(SJSDM_MOJO_BACKEND = "0")    # force PyTorch
```

Once reticulate has started Python, a later `Sys.setenv()` call is not copied into Python's `os.environ`. Restart R after changing it, or deliberately update the active Python process:

```{r mojo-switch-live}
set_sjsdm_backend <- function(backend = c("auto", "mojo", "torch")) {
  backend <- match.arg(backend)
  value <- c(auto = "auto", mojo = "1", torch = "0")[[backend]]
  reticulate::py_run_string(sprintf(
    "import os; os.environ['SJSDM_MOJO_BACKEND'] = '%s'", value
  ))
  invisible(value)
}

set_sjsdm_backend("auto")
```

Auto mode selects Mojo only for CPU float32 logit/probit operations when the worker exists and the complete response matrix has no missing values. CPU float64 and operations with missing responses use PyTorch for the whole operation. Forced Mojo mode remains strict and reports unsupported inputs.

## When does Mojo help?

Measured on v0.2.1 (2026-08-28, `-O 3` builds, clean workers):

- Faster in all 27 cells of the shape grid, from 1.69x to 3.14x.
- About 1.75x end-to-end at 10,000 sites x 200 species, and 1.50x at 3,000 sites x 60 species.
- Standard errors (`getSe`) always use PyTorch regardless of the setting, because Hessians need second-order autodiff.

The v0.2.0 release reported 1.9x and 1.3x on those same two end-to-end workloads (`Code/dev/bench_realdata.R`, `step_size = 10`). Those figures come from a separate benchmarking run and are kept labelled by release rather than overwritten.

Two caveats worth knowing:

1.  Cross-backend results are **statistically equivalent, not bit-identical**: float32 arithmetic differs slightly between the Mojo kernel and PyTorch's einsum, so loss histories agree only to \~$10^{-3}$ relative.
2.  Keep predictors **standardized**. sjSDM fits can diverge on raw-scale covariates (e.g., temperatures near 6000); this affects both backends. `scale(env)` fixes it.

The persistent worker retains each buffer's high-water capacity so repeated fits do not reallocate. A single unusually large fit can therefore keep that memory until the worker or R session exits. Restart the R session to release it; Python developers can call `sjSDM_py.mojo_bridge._WORKER.close()`.

# Data: simulate a community with known truth

The whole point of simulating is that you know the answer. We build a 200-site, 10-species community in which every ingredient of the model in the concepts section is set by hand, so that the next section can check each one against what sjSDM recovers.

The associations are given a **two-guild block structure** rather than a single constant value: species 1 to 5 are positively associated with each other, species 6 to 10 likewise, and the two guilds are negatively associated across the divide. A constant off-diagonal would make the recovery check vacuous, because there would be no pattern to recover -- only a single number to hit.

```{r simulate-data}
library(mvtnorm)

set.seed(42)
n_sites   <- 200
n_species <- 10

# 1. Environment: three standardized predictors.
X <- matrix(rnorm(n_sites * 3), n_sites, 3)
colnames(X) <- c("temp", "precip", "elev")

# 2. Environmental responses: one intercept and three slopes per species.
beta0 <- rnorm(n_species)                            # intercepts
beta  <- matrix(rnorm(3 * n_species), 3, n_species)  # slopes
beta_true <- rbind(beta0, beta)                      # 4 x species, the truth
rownames(beta_true) <- c("(Intercept)", colnames(X))
colnames(beta_true) <- paste0("sp", 1:n_species)

# 3. Associations: two guilds, positive within and negative between.
guild <- rep(c("A", "B"), each = n_species / 2)
Sigma <- outer(guild, guild, function(a, b) ifelse(a == b, 0.6, -0.3))
diag(Sigma) <- 1
stopifnot(all(eigen(Sigma)$values > 0))   # must be a valid covariance

# 4. Assemble: linear predictor + correlated site-level residual, then flip coins.
logit_p <- cbind(1, X) %*% beta_true
Z <- rmvnorm(n_sites, sigma = Sigma)
Y <- matrix(rbinom(n_sites * n_species, 1, plogis(logit_p + Z)),
            n_sites, n_species)
colnames(Y) <- colnames(beta_true)

# Sanity check before fitting anything: prevalence should not be extreme.
round(colMeans(Y), 3)
```

Note that step 4 is the generative model from the concepts section, written out literally: a linear predictor, plus one draw of $\mathbf{e}_i$ per site from $\mathcal{N}(\mathbf{0}, \boldsymbol{\Sigma})$, pushed through the inverse logit, then a Bernoulli trial. If you can read that block, you can read the model.

`Sigma` is positive definite and therefore full rank (rank 10 here). The fit below deliberately uses `df = 5`, so its association component is a rank-5 approximation to the truth rather than a correctly specified full-rank covariance model. That mirrors the regularized low-rank fits used in larger applications, but it also means the recovery check should preserve the guild pattern rather than reproduce every correlation magnitude.

Alternatively, use sjSDM's own simulator. The predictor matrix is `env_weights`, the response matrix is `response`, and the truth includes `species_weights` and `correlation`:

```{r simulate-SDM, eval=FALSE}
sim <- simulate_SDM(sites = 500, species = 20, correlation = TRUE)
str(sim)
```

Or analyze one of the bundled empirical datasets (e.g. `butterflies`, `eucalypts` -- presence/absence matrices shipped with the package).

# Fitting a basic model

The core function is `sjSDM()`. The three model components map directly to arguments:

```{r fit-basic}
fit <- sjSDM(
  Y = Y,                                        # sites x species 0/1 matrix
  env = linear(data = X, formula = ~ temp + precip + elev),
  biotic = bioticStruct(df = 5),                # rank-5 approximation to rank-10 truth
  family = binomial("logit"),
  iter = 100,                                   # training epochs
  sampling = 100,                               # MC samples per batch
  seed = 1,
  verbose = FALSE
)

summary(fit)
```

`summary()` prints three blocks: the log-likelihood and regularisation loss, the estimated species-species correlation matrix, and the coefficient table. With `se = TRUE` or after `getSe()`, the coefficient table gains standard errors, z values, and p values.

Notes on the knobs:

- `iter`, `sampling`, and `step_size` are the main speed/accuracy trade-offs. If omitted, `step_size` is 10% of the sites. More `sampling` reduces Monte Carlo noise in the likelihood.
- Select a link with `family = binomial("logit")` or `family = binomial("probit")`; Mojo handles both.
- Optimizer, learning-rate scheduling, and early stopping live in `sjSDMControl()` -- see `?sjSDMControl`.
- For non-linear responses, swap `linear(...)` for `DNN(data = X, ...)`.
- Regularize the association structure with, for example, `bioticStruct(df = 5, lambda = 1e-4, on_diag = TRUE)`; tune `lambda` by cross-validation.
- `df` is the rank $d$ from the concepts section. It caps how much association structure the model can represent.

# Does the model recover the truth?

This is the section to run before trusting a jSDM on real data, and the one to re-run whenever you change the fitting settings. We compare three things against the values used to simulate: the environmental coefficients, the species associations, and the fitted occurrence probabilities. Each comparison gets a table and a figure, because they answer different questions: the table says how close, the figure says in what way it is wrong.

## A reusable recovery summary

Every comparison below regresses estimates on truth. The slope is the informative number: 1 means unbiased, less than 1 means the estimates are shrunk toward zero, and the correlation says how well the *pattern* is recovered independently of scale.

```{r recovery-helper}
recovery_stats <- function(truth, est) {
  truth <- as.vector(truth); est <- as.vector(est)
  f <- stats::lm(est ~ truth)
  c(n           = length(truth),
    correlation = stats::cor(truth, est),
    slope       = unname(stats::coef(f)[2]),
    intercept   = unname(stats::coef(f)[1]),
    RMSE        = sqrt(mean((est - truth)^2)),
    bias        = mean(est - truth))
}
```

## Environmental coefficients

`coef(fit)[[1]]` is a species-by-coefficient matrix, so transposing it lines it up with `beta_true`. The row names come from `fit$names`.

```{r recovery-beta}
beta_hat <- t(coef(fit)[[1]])
dimnames(beta_hat) <- dimnames(beta_true)

tab_beta <- rbind(
  `all coefficients` = recovery_stats(beta_true, beta_hat),
  t(sapply(rownames(beta_true),
           function(r) recovery_stats(beta_true[r, ], beta_hat[r, ])))
)

knitr::kable(tab_beta, digits = 3,
             caption = "Recovery of environmental coefficients")
```

```{r recovery-beta-plot, fig.width=8, fig.height=4}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.6, 1))
cols <- grDevices::hcl.colors(nrow(beta_true), "Dark 3")

plot(beta_true, beta_hat, pch = 19, col = cols[row(beta_true)], asp = 1,
     xlab = "true coefficient", ylab = "estimated coefficient",
     main = "Environmental coefficients")
abline(0, 1, lty = 2)                                     # perfect recovery
abline(stats::lm(as.vector(beta_hat) ~ as.vector(beta_true)), col = "grey40")
legend("topleft", rownames(beta_true), col = cols, pch = 19,
       bty = "n", cex = 0.75)

boxplot(t(beta_hat - beta_true), col = "grey90",
        ylab = "estimate - truth", main = "Error by term")
abline(h = 0, lty = 2)
par(op)
```

With the settings above you should see a correlation near 0.96 and a slope near 1.0: the environmental part of a jSDM is estimated well and essentially without bias. The boxplot should straddle zero for every term. The intercept is usually the best-recovered coefficient, because every site contributes to it.

## Species associations

Two adjustments matter here, both of them consequences of points made in the concepts section.

- Compare **correlations**, not covariances. The latent scale is unidentified, so `getCov()` diagonals are not 1 and its off-diagonals are not on the same scale as `Sigma`.
- Report within-guild and between-guild pairs separately. Within each group the true correlation is constant, so a correlation coefficient is undefined there; the informative summary is the mean estimate and how far it is shrunk.

```{r recovery-assoc}
cor_true <- stats::cov2cor(Sigma)
cor_hat  <- getCor(fit)
dimnames(cor_true) <- dimnames(cor_hat) <-
  list(colnames(Y), colnames(Y))

ut   <- upper.tri(cor_true)                 # each pair once
same <- outer(guild, guild, "==")[ut]       # within-guild pairs

knitr::kable(rbind(`all pairs` = recovery_stats(cor_true[ut], cor_hat[ut])),
             digits = 3, caption = "Recovery of species associations, all pairs")

by_group <- function(keep, label) data.frame(
  pairs        = label,
  n            = sum(keep),
  true         = mean(cor_true[ut][keep]),
  estimated    = mean(cor_hat[ut][keep]),
  sd_estimated = stats::sd(cor_hat[ut][keep]),
  shrinkage    = mean(cor_hat[ut][keep]) / mean(cor_true[ut][keep]),
  correct_sign = mean(sign(cor_hat[ut][keep]) == sign(cor_true[ut][keep]))
)

knitr::kable(rbind(by_group(same,  "within guild (true +0.6)"),
                   by_group(!same, "between guild (true -0.3)")),
             digits = 3, row.names = FALSE,
             caption = "Estimated associations by guild pair")
```

```{r recovery-assoc-plot, fig.width=9, fig.height=3.6}
op <- par(mfrow = c(1, 3), mar = c(4, 4, 2.6, 1))

heat <- function(M, main) {
  image(seq_len(ncol(M)), seq_len(nrow(M)), t(M[nrow(M):1, ]),
        col = grDevices::hcl.colors(50, "Blue-Red 3"),
        breaks = seq(-1, 1, length.out = 51),
        axes = FALSE, xlab = "", ylab = "", main = main)
  axis(1, seq_len(ncol(M)), colnames(M), las = 2, cex.axis = 0.6, tick = FALSE)
  axis(2, seq_len(nrow(M)), rev(rownames(M)), las = 2, cex.axis = 0.6, tick = FALSE)
}
heat(cor_true, "True correlations")
heat(cor_hat,  "Estimated correlations")

plot(cor_true[ut], cor_hat[ut], pch = 19, asp = 1,
     col = ifelse(same, "#1b7837", "#762a83"),
     xlim = c(-1, 1), ylim = c(-1, 1),
     xlab = "true correlation", ylab = "estimated correlation",
     main = "Pairwise")
abline(0, 1, lty = 2); abline(h = 0, col = "grey80"); abline(v = 0, col = "grey80")
abline(stats::lm(cor_hat[ut] ~ cor_true[ut]), col = "grey40")
legend("topleft", c("within guild", "between guild"),
       col = c("#1b7837", "#762a83"), pch = 19, bty = "n", cex = 0.75)
par(op)
```

Read the three panels together. The two heatmaps should show the same block structure, the estimated one noisier and paler. The scatter should show the fitted line clearly **below** the 1:1 dashed line: with these settings the mean within-guild correlation comes back around 0.25 against a truth of 0.6, and the between-guild mean around -0.16 against -0.3. The correlation across all pairs is about 0.72 and the slope about 0.46.

That gap is expected in this deliberately low-rank teaching fit, and it is worth understanding because several sources of attenuation recur in real analyses.

- The rank-5 fit can represent only five covariance directions from the full-rank truth, so rank misspecification attenuates some associations before sampling and optimisation effects enter.
- Binary data carry little information per observation. Each 0/1 is a single coin flip, so the latent correlation is only weakly constrained.
- The rank-$d$ factorisation plus any `lambda` penalty is a shrinkage estimator by construction.
- Monte Carlo noise in the likelihood adds variance that biases estimated correlations toward zero.
- Finite `iter` means the optimiser may not have fully converged.

The practical rule that follows: **trust the sign and the ordering, not the magnitude.** Comparisons between datasets fitted with identical settings are meaningful; a single correlation read off as a calibrated effect size is not.

## Occurrence probabilities

The environmental and association parts can each be imperfect while predictions are still good, so check the prediction directly.

```{r recovery-prob, fig.width=5, fig.height=4.6}
p_true <- plogis(logit_p)     # truth conditional on the environment only
p_hat  <- predict(fit)

knitr::kable(rbind(`occurrence probability` = recovery_stats(p_true, p_hat)),
             digits = 3, caption = "Recovery of occurrence probabilities")

par(mar = c(4.2, 4.2, 2.6, 1))
smoothScatter(as.vector(p_true), as.vector(p_hat), nrpoints = 0, asp = 1,
              xlab = "true P(occurrence)", ylab = "predicted P(occurrence)",
              main = "Site x species probabilities")
abline(0, 1, lty = 2)
```

Expect a correlation around 0.98 and a slope near 1. Prediction is the easiest of the three targets, which is why good predictive performance is weak evidence that the association estimates are trustworthy. Report recovery of the quantity you intend to interpret, not the one that looks best.

## What to do when recovery is poor

If you run this section on a simulation resembling your real data and the coefficients do not come back, fix that before touching real data. In order of how often each is the culprit: unstandardized predictors, too few sites relative to species, extreme prevalence, `iter` too small for convergence, `sampling` too small so the likelihood is too noisy, and `df` too small to represent the association structure you simulated.

# Inspecting species associations

```{r associations}
cov_mat <- getCov(fit)      # estimated species covariance matrix
round(cov_mat, 2)

getCor(fit)                 # covariance rescaled to correlations

# The diagonal is NOT 1: the latent scale is free. Use getCor() for
# anything you intend to compare against another model or a truth.
round(diag(cov_mat), 2)
```

Species weights (the actual fitted parameter matrices) are available via `getWeights(fit)` / `setWeights()` if you need raw access.

# Detecting assembly processes

The three subsections below each simulate two communities that differ in exactly one process, fit the same model to both, and ask whether the model's output separates them. This is the design to copy when you want to know whether your own dataset can support a claim: build the contrast you care about, and check that the model can see it before you go looking for it in real data.

All three read `summary(anova(m))`, whose rows are the partition components and whose `R2 Nagelkerke` column is the one to compare. Note the `setNames()` call: extracting a column from that data frame drops the component names, so they have to be reattached before the results can be stacked or plotted.

```{r partition-helper}
partition_of <- function(m) {
  av <- summary(anova(m))
  stats::setNames(av[, "R2 Nagelkerke"], rownames(av))
}

partition_barplot <- function(tab, components, main, labels = rownames(tab)) {
  m <- t(tab[, components, drop = FALSE])
  barplot(m, beside = TRUE, names.arg = labels,
          col = grDevices::hcl.colors(length(components), "Dark 3"),
          ylab = "R2 (Nagelkerke)", main = main,
          ylim = c(0, max(m) * 1.35),       # headroom so the legend clears the bars
          legend.text = components,
          args.legend = list(x = "topleft", bty = "n", cex = 0.85,
                             horiz = length(components) <= 2))
}
```

## Do fits recover known association strength?

Simulate two communities that differ **only** in the strength of their true species associations, fit the same environmental model to each, and verify that both the estimated correlations and the variation partition track the truth.

```{r assoc-recovery}
simulate_community <- function(assoc_strength, seed = 7) {
  set.seed(seed)  # common random numbers isolate the covariance change
  Sig <- matrix(assoc_strength, n_species, n_species)
  diag(Sig) <- 1
  Zc <- mvtnorm::rmvnorm(n_sites, sigma = Sig)
  matrix(rbinom(n_sites * n_species, 1, plogis(logit_p + Zc)),
         n_sites, n_species)
}

fit_community <- function(Y_comm, seed = 1) {
  sjSDM(Y = Y_comm, env = linear(data = X),
        biotic = bioticStruct(df = 5), family = binomial("logit"),
        iter = 100, sampling = 100, seed = seed, verbose = FALSE)
}

mean_offdiag <- function(m) {
  C <- getCor(m)
  mean(C[upper.tri(C)])
}

m_weak   <- fit_community(simulate_community(0.1))
m_strong <- fit_community(simulate_community(0.6))

assoc_tab <- rbind(`weak (true r = 0.1)`   = partition_of(m_weak),
                   `strong (true r = 0.6)` = partition_of(m_strong))
assoc_tab <- cbind(assoc_tab,
                   `true correlation`      = c(0.1, 0.6),
                   `mean est. correlation` = c(mean_offdiag(m_weak),
                                               mean_offdiag(m_strong)))

knitr::kable(assoc_tab, digits = 3,
             caption = "Association strength: truth, partition, and estimate")
```

```{r assoc-recovery-plot, fig.width=8, fig.height=3.8}
op <- par(mfrow = c(1, 2), mar = c(4, 4.2, 2.6, 1))
partition_barplot(assoc_tab, c("Abiotic", "Associations"),
                  "Variation partitioning")
plot(assoc_tab[, "true correlation"], assoc_tab[, "mean est. correlation"],
     pch = 19, cex = 1.4, xlim = c(0, 0.8), ylim = c(0, 0.8), asp = 1,
     xlab = "true mean correlation", ylab = "estimated mean correlation",
     main = "Association strength")
abline(0, 1, lty = 2)
text(assoc_tab[, "true correlation"], assoc_tab[, "mean est. correlation"],
     labels = c("weak", "strong"), pos = 3, cex = 0.8)
par(op)
```

The strong-association community should show a substantially larger `Associations` component and a larger mean fitted correlation, while `Abiotic` stays essentially unchanged -- the environmental signal is identical by construction. Both points should sit below the 1:1 line, which is the same shrinkage seen in the recovery section, and the reason the useful comparison is between the two points rather than between either point and the truth.

## Do fits detect dispersal limitation?

Dispersal limitation leaves a characteristic fingerprint: species occur (or fail to) in **spatially coherent patches** that no measured environmental variable explains. We simulate this by drawing latent site effects from a *spatially autocorrelated* Gaussian field over the coordinates -- weak or strong -- while keeping environmental responses identical.

This section asks two questions, and the second is the more important one. Does a spatial sjSDM assign more deviance to the spatial component when dispersal limitation is strong? And what happens to that signal if you leave the spatial component out?

**A warning about the size of the spatial basis.** `generateSpatialEV()` returns a large set of Moran eigenvector maps, 63 columns for the 200 sites used here. Fitting all of them saturates the spatial component: it explains almost everything under both weak and strong dispersal, and the contrast disappears entirely. That is overfitting, not detection. Use a modest number of leading eigenvectors, or select them by cross-validation.

```{r dispersal}
# Exponentially decaying spatial covariance over coordinates:
spatial_field <- function(coords, range = 0.5) {
  D <- as.matrix(dist(coords))
  mvtnorm::rmvnorm(1, sigma = exp(-D / range))   # one site-level draw
}

simulate_dispersal_community <- function(dispersal, coords, seed = 1) {
  set.seed(seed)
  # Latent patchiness shared by all species, scaled by dispersal strength.
  # Everything else, including beta, is held fixed.
  field <- drop(spatial_field(coords))
  Zd <- dispersal * outer(field, rep(1, n_species))
  matrix(rbinom(n_sites * n_species, 1, plogis(logit_p + Zd)),
         n_sites, n_species)
}

coords <- matrix(runif(n_sites * 2), n_sites, 2)
SP     <- generateSpatialEV(coords)
dim(SP)                       # 63 eigenvectors: too many to fit blindly
SP10 <- SP[, 1:10]            # leading MEMs only

fit_with_space <- function(Y_comm) {
  sjSDM(Y = Y_comm, env = linear(data = X),
        spatial = linear(data = SP10, formula = ~ 0 + .),
        biotic = bioticStruct(df = 5), family = binomial("logit"),
        iter = 100, seed = 1, verbose = FALSE)
}

fit_without_space <- function(Y_comm) {
  sjSDM(Y = Y_comm, env = linear(data = X),
        biotic = bioticStruct(df = 5), family = binomial("logit"),
        iter = 100, sampling = 100, seed = 1, verbose = FALSE)
}

Y_disp <- lapply(c(weak = 0.2, strong = 2.0),
                 function(d) simulate_dispersal_community(d, coords))

disp_tab <- rbind(
  `weak dispersal limitation`   = partition_of(fit_with_space(Y_disp$weak)),
  `strong dispersal limitation` = partition_of(fit_with_space(Y_disp$strong))
)

knitr::kable(disp_tab, digits = 3,
             caption = "Dispersal limitation, spatial component fitted")
```

The spatial component grows markedly under strong dispersal limitation, from roughly 0.39 to 0.81. Two other numbers matter as much. The environmental component barely moves, from about 0.92 to 0.84, because the true `beta` was held fixed. And the association component stays essentially flat, near 0.19 in both rows: with somewhere legitimate to put the spatial signal, the model puts it there and leaves the residual associations alone.

Now drop the spatial term and refit exactly the same data.

```{r dispersal-nospace}
m_ns <- lapply(Y_disp, fit_without_space)

mean_abs_cor <- function(m) {
  C <- getCor(m)
  mean(abs(C[upper.tri(C)]))
}

nospace_tab <- cbind(
  t(sapply(m_ns, partition_of)),
  `mean |correlation|` = sapply(m_ns, mean_abs_cor)
)
rownames(nospace_tab) <- c("weak, no spatial term", "strong, no spatial term")

knitr::kable(nospace_tab, digits = 3,
             caption = "Same data, spatial component omitted")
```

```{r dispersal-plot, fig.width=8, fig.height=3.8}
op <- par(mfrow = c(1, 2), mar = c(4, 4.2, 2.6, 1))
partition_barplot(disp_tab, c("Abiotic", "Spatial", "Associations"),
                  "Spatial component fitted", labels = c("weak", "strong"))

assoc_cmp <- rbind(
  weak   = c(`with spatial term` = disp_tab[1, "Associations"],
             `without`           = nospace_tab[1, "Associations"]),
  strong = c(`with spatial term` = disp_tab[2, "Associations"],
             `without`           = nospace_tab[2, "Associations"])
)
partition_barplot(assoc_cmp, colnames(assoc_cmp),
                  "Association component only")
par(op)
```

This is the point of the section. With no spatial term the dispersal signal does not disappear; it moves into the residual associations. `Associations` rises from about 0.15 to about 0.58, and the mean absolute species correlation rises from about 0.16 to about 0.55. Compare that with the flat 0.19 in the previous table, on exactly the same data, from a model that was given somewhere else to put the signal.

Nothing generated those correlations except a smooth spatial field shared by all species. There is no interaction anywhere in the simulation.

An analyst who fitted only the non-spatial model, saw a mean absolute correlation of 0.55, and concluded that the community shows strong biotic structuring would be wrong, and nothing in the model output would say so. Fit space whenever you have coordinates, even when space is not your question.

## Do fits detect environmental filtering?

Here, stronger environmental filtering means occurrence responds more strongly to the measured gradients. We hold everything else fixed and scale the true environmental coefficients, so the truth changes in exactly one direction.

```{r filtering}
simulate_filtered_community <- function(beta_scale, seed = 1) {
  set.seed(seed)
  Sig <- matrix(0.1, n_species, n_species); diag(Sig) <- 1
  lp <- cbind(1, X) %*% rbind(beta0, beta_scale * beta)
  Zf <- mvtnorm::rmvnorm(n_sites, sigma = Sig)
  matrix(rbinom(n_sites * n_species, 1, plogis(lp + Zf)),
         n_sites, n_species)
}

fit_plain <- function(Y_comm) {
  sjSDM(Y = Y_comm, env = linear(data = X),
        biotic = bioticStruct(df = 5), family = binomial("logit"),
        iter = 100, sampling = 100, seed = 1, verbose = FALSE)
}

filt_scales <- c(0.25, 1.0, 2.0)
filt_tab <- t(sapply(filt_scales,
                     function(s) partition_of(fit_plain(simulate_filtered_community(s)))))
rownames(filt_tab) <- paste0("beta x ", filt_scales)
filt_tab <- cbind(filt_tab, `true beta scale` = filt_scales)

knitr::kable(filt_tab, digits = 3,
             caption = "Environmental filtering: variation partitioning")
```

```{r filtering-plot, fig.width=8, fig.height=3.8}
op <- par(mfrow = c(1, 2), mar = c(4, 4.2, 2.6, 1))
partition_barplot(filt_tab, c("Abiotic", "Associations"),
                  "Increasing environmental filtering")
matplot(filt_scales, filt_tab[, c("Abiotic", "Associations", "Full")],
        type = "b", pch = 19, lty = 1, log = "x",
        col = grDevices::hcl.colors(3, "Dark 3"),
        xlab = "true coefficient scale", ylab = "R2 (Nagelkerke)",
        main = "Partition vs filtering strength")
legend("bottomright", c("Abiotic", "Associations", "Full"),
       col = grDevices::hcl.colors(3, "Dark 3"), lty = 1, pch = 19,
       bty = "n", cex = 0.8)
par(op)
```

Under strong filtering you should see both a larger total $R^2$ and a partition dominated by the environment-only component. When filtering is weak the residual association component absorbs more of the unexplained structure, which is the same conflation problem stated from the other direction.

## What these tests cannot tell you

Each contrast above changes one thing and shows the model responds to it. That is a *necessary* condition for inference, not a sufficient one. Three limits are worth stating plainly.

- **The tests are calibrated on data generated by the model being fitted.** Real data are not, and recovery on real data will be worse than anything shown here.
- **They compare two known conditions.** In a real dataset you have one condition and no truth, so you can report that the association component is large but not that it is larger than it should be.
- **They do not identify the cause.** The dispersal section deliberately produces a strong `Associations` signal from a process that is not an interaction. Nothing in the output distinguishes the two; only the design of the covariate set does.

# Prediction and model fit

```{r predict}
pred <- predict(fit)        # sites x species occurrence probabilities
head(pred[, 1:4])

# Predict at new environmental values:
newdata <- data.frame(temp = c(-1, 0, 1), precip = 0, elev = 0)
predict(fit, newdata = newdata)

Rsquared(fit)               # Nagelkerke-style R^2 for the joint model
```

Interpret this $R^2$ as a relative measure across models fitted to the same data, not as an absolute goodness-of-fit. Binary responses cap the achievable value well below 1, and the simulated data above give roughly 0.2 even though the model is correctly specified.

# Variation partitioning (ANOVA)

`anova()` refits reduced models (environment-only, associations-only, and space-only when a spatial component is present) and decomposes explained deviance. This is slower than fitting the full model because of the refits.

```{r anova}
av <- anova(fit)
av
summary(av)                 # includes Nagelkerke R^2 per component
plot(av)                    # barplot of the partition
```

Reading the partition:

- **Abiotic** is deviance explained by the environmental predictors.
- **Spatial** appears only when the model has a spatial component, and is deviance explained by the spatial eigenvectors.
- **Associations** is deviance explained by the residual covariance, that is, structure the other components missed.
- **Full** is the joint model. Components do not sum to `Full`, because they share explained deviance; that shared fraction is the point of partitioning.

A large `Associations` value is a prompt to ask what covariate is missing, not a finding about interactions.

For a spatial model, `internalStructure()` further splits the association component into local and spatially structured parts, and `plotAssemblyEffects()` visualizes species responses along gradients.

# Cross-validation

Use `sjSDM_cv()` to tune regularization strength. Two important details:

- It takes **raw data**, not a fitted model.
- The tuning argument is `tune_steps`, not `IT`.

```{r cv, eval=FALSE}
options(sjSDM.dev = here::here("sjSDM"))

cv <- sjSDM_cv(
  Y = Y, env = linear(data = X),
  biotic = bioticStruct(df = 5),
  family = binomial("logit"),
  tune = "random", tune_steps = 5,
  CV = 3, iter = 100, sampling = 100,
  n_cores = 2
)

plot(cv)                            # CV curves over lambda
```

Parallel workers load sjSDM themselves; when developing against this repo, point workers at the source checkout with `options(sjSDM.dev = here::here("sjSDM"))`.

# Spatial models

Add spatial predictors (Moran's eigenvector maps from site coordinates):

```{r spatial}
coords <- matrix(runif(n_sites * 2), n_sites, 2)
SP <- generateSpatialEV(coords)          # spatial eigenvectors

fit_sp <- sjSDM(
  Y = Y,
  env = linear(data = X),
  spatial = linear(data = SP, formula = ~ 0 + .),
  family = binomial("logit"),
  iter = 100, seed = 1, verbose = FALSE
)

anova(fit_sp)   # partitions deviance into environment, space, associations
```

Fit the spatial component whenever you have coordinates, even if space is not your question. Omitting it does not remove spatial structure from the data; it moves that structure into the residual associations, where it is easy to misread as biotic signal.

# Standard errors and plots

`plot(fit)` draws coefficient estimates with error bars. Computing the error bars requires per-species Hessians, which is the most expensive post-fitting step -- request SEs once with `se = TRUE`, or let `plot()` trigger `getSe()`.

```{r se-plot, eval=FALSE}
fit_se <- sjSDM(Y = Y, env = linear(data = X),
                biotic = bioticStruct(df = 5),
                family = binomial("logit"),
                se = FALSE, iter = 100, seed = 1, verbose = FALSE)

# getSe() returns an updated model; the matrix is stored in $se.
fit_se <- getSe(fit_se)
se <- fit_se$se

p_coef <- plot(fit_se)      # ggplot2 coefficient plot
p_coef
```

Standard errors always use PyTorch, because the Hessian needs second-order automatic differentiation that the Mojo kernel does not provide.

Assembly-effect diagnostics for a spatial ANOVA:

```{r assembly, eval=FALSE}
intstr <- internalStructure(anova(fit_sp))
plotAssemblyEffects(intstr)
```

# Benchmarking torch vs. Mojo yourself

A quick timing harness compares backends on the same data and model seed. Because Python is already running, it uses the `set_sjsdm_backend()` helper defined earlier rather than `Sys.setenv()`:

```{r bench}
bench_fit <- function(backend) {
  set_sjsdm_backend(backend)
  t0 <- Sys.time()
  m <- sjSDM(Y = Y, env = linear(data = X),
             biotic = bioticStruct(df = 5),
             family = binomial("logit"),
             iter = 100, sampling = 100, seed = 123, verbose = FALSE)
  c(seconds = as.numeric(difftime(Sys.time(), t0, units = "secs")),
    final_loss = tail(m$history, 1))
}

bench_fit("torch")
bench_fit("auto")
set_sjsdm_backend("auto")
```

Run each backend more than once for a serious benchmark and report the distribution of timings. Losses should be close, not bit-identical, because the backends use different random-number generation and float32 reduction order.

# Troubleshooting

- **`is_torch_available()` is FALSE.** Run `reticulate::py_config()`. Most often `RETICULATE_PYTHON` points nowhere or was set after Python started. Set it before loading sjSDM and restart R.
- **Mojo backend seems ignored.** Check `file.exists(here::here("work/mojo-backend/mc_grad_server_bin"))`. If missing, redo Step 3. If Python is already active, switch with `set_sjsdm_backend()` or restart R.
- **Fit diverges to a constant loss.** Unstandardized predictors. Use `scale(env)`. Affects both backends equally.
- **`torch.randn` dimension errors.** Fixed in this repository; make sure workers load the patched code (`options(sjSDM.dev = ...)` in parallel CV).
- **Want GPU (Apple MPS)?** `options(sjSDM.device = "mps")`; note the Mojo backend is CPU-only, so MPS fits fall back to torch.
- **pixi or build step fails.** Confirm you are inside `work/mojo-backend/` when running `pixi` commands; `pixi run mojo` must resolve the Mojo standard library through the pixi environment.
- **Partition columns come out unnamed.** Extracting a column from `summary(anova(m))` drops the component names. Use `setNames(av[, "R2 Nagelkerke"], rownames(av))`, as in the `partition_of()` helper above.
- **Estimated correlations are much smaller than expected.** This is shrinkage, not a failure; see the association recovery section. Compare across models rather than against an absolute value.

# Where to go next

- Package docs: `?sjSDM`, `?bioticStruct`, `?linear`, `?sjSDMControl`, `?sjSDM_cv`.
- Benchmarks and parity checks: `Code/dev/benchmarks/`, `Code/dev/bench_realdata.R`, `Code/dev/bench_real_occurrence.R`.
- Mojo kernel sources and validation harness: `work/mojo-backend/mojo/` and `Feasibility_study.md`.
- Upgrading Mojo/MAX versions: see the "Maintaining the Mojo/MAX pin" section of [Code/README.md](README.md).
- Original sjSDM paper: Pichler & Hartig (2021), *A new method for faster and more accurate joint species distribution modelling* (Methods in Ecology & Evolution).
- On the interpretation of residual associations: Poggiato et al. (2021), *On the interpretations of joint modeling in community ecology* (Trends in Ecology & Evolution); Blanchet et al. (2020), *Co-occurrence is not evidence of ecological interactions* (Ecology Letters).
- Alternative jSDM implementations worth knowing: HMSC (Bayesian, richer trait and phylogeny structure, slower) and boral (latent variable models with a similar factorisation).
