---
title: "sjSDM from R: a complete tutorial, including the Mojo CPU accelerator"
author: "sjSDM developers"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,   # heavy model fits; knit fast and let readers run chunks
  comment = "#>"
)
```

This tutorial covers:

1. What joint species distribution models (jSDMs) are and what **sjSDM** does.
2. Installing the Mojo-accelerated build -- *no Python knowledge required*.
3. Every major feature: fitting, species associations, prediction, $R^2$,
   variation partitioning (ANOVA), cross-validation, spatial models,
   standard errors, plots, and simulation.
4. The **Mojo CPU backend**: what it is and how to control it with one
   environment variable.

Chunks are set to `eval = FALSE` so the document renders instantly; run them
interactively top-to-bottom instead (total runtime of the full walkthrough is
a few minutes on a laptop).

# Background: what is sjSDM?

A *species distribution model* (SDM) relates species occurrence to the
environment. A *joint* SDM (jSDM) models all species at once and estimates
**species associations** -- residual correlations between species after
accounting for the environment. These associations can indicate biotic
interactions, shared unmeasured habitat variables, or dispersal history.

sjSDM implements scalable jSDMs using deep-learning machinery (PyTorch):

- Multivariate probit or logit likelihoods with a **low-rank covariance**
  for species associations (`bioticStruct(df = ...)`).
- Environmental (and optionally spatial) predictor layers via `linear()`
  formulas, plus optional deep neural network layers via `DNN()`.
- Fast minibatch-gradient-descent fitting that scales to tens of thousands
  of sites.

The **Mojo backend** replaces the inner Monte-Carlo likelihood computation
(the hot loop) with a hand-written kernel in the Mojo language, compiled to
native Apple Silicon code. On typical workloads it is roughly 1.3--3x faster
than PyTorch on CPU while producing statistically equivalent results.

# Installation

## Who this tutorial is for

This tutorial covers the **Mojo-accelerated CPU build** of sjSDM, which
currently requires **Apple Silicon (macOS, arm64)** and a full clone of the
development repository at `dougwyu/s-jSDM`.

> **Not on Apple Silicon, or just want plain sjSDM?** You don't need this
> tutorial. Install the original, pure-PyTorch sjSDM instead -- it is fully
> supported on macOS (Intel and Apple Silicon), Windows, and Linux:
>
> ```{r fallback, eval=FALSE}
> remotes::install_github("TheoreticalEcology/s-jSDM/sjSDM")
> library(sjSDM)
> install_sjSDM()   # downloads a self-contained Python environment; no Python skills needed
> ```
>
> Everything from Section 4 onward of this tutorial works identically with
> that install; just skip Sections 2--3 (Mojo setup and switching).

## One install path: clone + pixi

The Mojo backend lives alongside a pinned Python environment (PyTorch 2.5.1)
defined with [pixi](https://pixi.sh), a single-binary package manager. You
will install pixi once, let it build the environment, and compile the Mojo
worker binary. No Python knowledge is needed at any step.

**Step 1 -- clone the repository:**

```{bash clone}
# git clone https://github.com/dougwyu/s-jSDM.git
# cd s-jSDM
```

**Step 2 -- install pixi** (one-time; see https://pixi.sh for alternatives):

```{bash pixi}
# macOS/Linux:
# curl -fsSL https://pixi.sh/install.sh | sh
```

**Step 3 -- create the pinned Python environment and build the Mojo worker.**
From the repository root:

```{bash build}
# cd work/mojo-backend
# pixi install
# pixi run mojo build -O 3 mojo/mc_logit_grad_server.mojo -o mc_grad_server_bin
```

The `-O 3` flag matters: unoptimized Mojo builds are roughly 50x slower.
When finished, `work/mojo-backend/mc_grad_server_bin` exists -- that is
the accelerated likelihood kernel sjSDM will find and use automatically.

**Step 4 -- point R at the environment and load the package.** Do this at
the start of every R session, **before** `library(sjSDM)`:

```{r reticulate}
Sys.setenv(RETICULATE_PYTHON = here::here(
  "work/mojo-backend/.pixi/envs/default/bin/python"
))
stopifnot(file.exists(Sys.getenv("RETICULATE_PYTHON")))

# library(sjSDM) works for normal use. When working from the repo checkout,
# devtools::load_all() runs the repo code (including the patched Python
# backend) instead of your installed library copy:
devtools::load_all(here::here("sjSDM"), quiet = TRUE)

is_torch_available()
install_diagnostic()   # detailed report if something is off
```

> **Key idea for R users:** R talks to Python through the `reticulate`
> package, which sjSDM handles internally. The only thing you ever do with
> Python is tell R where the environment lives via `RETICULATE_PYTHON`.
> The Mojo binary is a separate native program that sjSDM also finds on its
> own -- you never invoke it directly.

# Using the Mojo CPU backend

## What it is

The Mojo backend accelerates the Monte Carlo logit/probit likelihood and its
analytic gradients on **CPU**. It consists of the two pieces you built
during installation:

| Piece | File | Role |
|---|---|---|
| Persistent worker | `work/mojo-backend/mc_grad_server_bin` | Long-lived process computing loss + gradients per training batch |
| R-side bridge | `sjSDM/inst/python/sjSDM_py/mojo_bridge.py` | Feeds tensors/seeds to the worker through a pipe |

You do **not** call either directly. You control the backend with one
environment variable.

## Turning it on / off

```{r mojo-switch}
# AUTO (default behavior): Mojo is used automatically whenever the worker
# binary exists AND the fit runs on CPU with a logit/probit link.
# Set explicitly with:
Sys.setenv(SJSDM_MOJO_BACKEND = "auto")   # default; silent fallback to torch

# Force Mojo on (errors if the binary is missing):
Sys.setenv(SJSDM_MOJO_BACKEND = "1")

# Force the pure-PyTorch path:
Sys.setenv(SJSDM_MOJO_BACKEND = "0")
```

Set these **before** calling `sjSDM()` (or restart R after changing them).
In `"auto"` mode there is nothing to configure: if the binary is present you
get the speedup; if anything disqualifies a fit (GPU device, float64, NaN
responses, Gaussian link), sjSDM silently uses PyTorch for those batches.

## When does Mojo help?

Benchmarks in this repo (`Code/bench_realdata.R`, `Code/benchmarks/`) show:

- Larger problems win most: ~1.9x end-to-end at 10,000 sites x 200 species;
  ~1.3x at 3,000 x 60; up to ~3.5x per-batch in shape-grid tests.
- Small runs are still fine -- the backend wins essentially everywhere on
  CPU as of the latest SIMD kernels.
- Standard errors (`getSe`) always use PyTorch regardless of the setting,
  because Hessians need second-order autodiff.

Two caveats worth knowing:

1. Cross-backend results are **statistically equivalent, not bit-identical**:
   float32 arithmetic differs slightly between the Mojo kernel and PyTorch's
   einsum, so loss histories agree only to ~$10^{-3}$ relative.
2. Keep predictors **standardized**. sjSDM fits can diverge on raw-scale
   covariates (e.g., temperatures near 6000); this affects both backends.
   `scale(env)` fixes it.

# Data: simulate a community

We create a 200-site, 10-species community with known environmental
responses and known species associations, so we can check that sjSDM
recovers them.

```{r simulate-data}
set.seed(42)
n_sites   <- 200
n_species <- 10

X <- matrix(rnorm(n_sites * 3), n_sites, 3)
colnames(X) <- c("temp", "precip", "elev")

beta  <- matrix(rnorm(3 * n_species), 3, n_species)     # env responses
Sigma <- matrix(0.5, n_species, n_species); diag(Sigma) <- 1  # associations

library(mvtnorm)
logit_p <- cbind(1, X) %*% rbind(rnorm(n_species), beta)
Z <- rmvnorm(n_sites, sigma = Sigma)
Y <- matrix(rbinom(n_sites * n_species, 1, plogis(logit_p + Z)),
            n_sites, n_species)
```

Alternatively, use sjSDM's own simulator (returns a list with `env`,
`species`, `response`, and truth):

```{r simulate-SDM, eval=FALSE}
sim <- simulate_SDM(n_sites = 500, n_species = 20, correlation = TRUE)
str(sim)
```

Or analyze one of the bundled empirical datasets (e.g. `butterflies`,
`eucalypts` -- presence/absence matrices shipped with the package).

# Fitting a basic model

The core function is `sjSDM()`. The three model components map directly to
arguments:

```{r fit-basic}
fit <- sjSDM(
  Y = Y,                                        # sites x species 0/1 matrix
  env = linear(data = X, formula = ~ temp + precip + elev),
  biotic = bioticStruct(df = 5),                # rank-5 covariance factor
  iter = 100,                                   # gradient steps
  sampling = 100,                               # MC samples per step
  verbose = FALSE
)

summary(fit)
```

Notes on the knobs:

- `iter`, `sampling`, `batch_size` (set internally to ~10% of sites) are the
  main speed/accuracy trade-offs. More `sampling` reduces Monte Carlo noise
  in the likelihood.
- `link = "probit"` is also supported; the Mojo backend handles both.
- Optimizer, learning-rate scheduling, and early stopping live in
  `sjSDMControl()` -- see `?sjSDMControl`.
- For non-linear responses, swap `linear(...)` for `DNN(data = X, ...)`.
- Regularize with e.g. `bioticStruct(df = 5, on_diag = 1e-4)` penalties;
  tune them by cross-validation (Section 8).

# Inspecting species associations

```{r associations}
cov_mat <- getCov(fit)      # estimated species covariance matrix
round(cov_mat, 2)

getCor(fit)                 # same, as correlations

# Do we recover the strong positive association we simulated?
mean(cov_mat[upper.tri(cov_mat)])
```

Species weights (the actual fitted parameter matrices) are available via
`getWeights(fit)` / `setWeights()` if you need raw access.

# Prediction and model fit

```{r predict}
pred <- predict(fit)        # sites x species occurrence probabilities
head(pred[, 1:4])

# Predict at new environmental values:
newdata <- data.frame(temp = c(-1, 0, 1), precip = 0, elev = 0)
predict(fit, newdata = newdata)

Rsquared(fit)               # Nagelkerke-style R^2 for the joint model
```

# Variation partitioning (ANOVA)

`anova()` refits reduced models (environment-only, associations-only) and
decomposes explained deviance. This is slower than fitting the full model
because of the refits.

```{r anova}
av <- anova(fit)
av
summary(av)                 # includes Nagelkerke R^2 per component
plot(av)                    # barplot of the partition
```

For a spatial model (Section 9), `internalStructure()` further splits the
association component into local vs. spatially structured parts, and
`plotAssemblyEffects()` visualizes species responses along gradients.

# Cross-validation

Use `sjSDM_cv()` to tune regularization strength. Two things people trip on:

- It takes **raw data**, not a fitted model.
- The tuning argument is `tune_steps`, not `IT`.

```{r cv}
cv <- sjSDM_cv(
  Y = Y,
  env = linear(data = X),
  biotic = bioticStruct(df = 5),
  tune = "random", tune_steps = 5,   # random search over penalty grid
  CV = 3,                            # 3-fold cross-validation
  iter = 100, sampling = 100,
  n_cores = 2                        # parallel folds
)
plot(cv)                            # CV curves over lambda
```

Parallel workers load sjSDM themselves; when developing against this repo,
point workers at the source checkout with
`options(sjSDM.dev = here::here("sjSDM"))`.

# Spatial models

Add spatial predictors (Moran's eigenvector maps from site coordinates):

```{r spatial}
coords <- matrix(runif(n_sites * 2), n_sites, 2)
SP <- generateSpatialEV(coords)          # spatial eigenvectors

fit_sp <- sjSDM(
  Y = Y,
  env = linear(data = X),
  spatial = linear(data = SP),
  iter = 100, verbose = FALSE
)

anova(fit_sp)   # partitions deviance: environment | space | associations
```

# Standard errors and plots

`plot(fit)` draws coefficient estimates with error bars. Computing the error
bars requires per-species Hessians, which is the most expensive post-fitting
step -- request SEs once with `se = TRUE`, or let `plot()` trigger
`getSe()`.

```{r se-plot}
fit_se <- sjSDM(Y = Y, env = linear(data = X),
                biotic = bioticStruct(df = 5),
                se = TRUE, iter = 100, verbose = FALSE)

se <- getSe(fit_se)         # standard errors for env coefficients

p_coef <- plot(fit_se)      # ggplot2 coefficient plot
p_coef
```

Assembly-effect diagnostics for a spatial ANOVA:

```{r assembly}
intstr <- internalStructure(anova(fit_sp))
plotAssemblyEffects(intstr)
```

# Benchmarking torch vs. Mojo yourself

A quick timing harness comparing backends on the same fixed-seed fit:

```{r bench}
bench_fit <- function(backend) {
  Sys.setenv(SJSDM_MOJO_BACKEND = backend)
  t0 <- Sys.time()
  m <- sjSDM(Y = Y, env = linear(data = X),
             biotic = bioticStruct(df = 5),
             iter = 100, sampling = 400, verbose = FALSE)
  c(seconds = as.numeric(difftime(Sys.time(), t0, units = "secs")),
    final_loss = tail(m$losses, 1))       # loss trajectories should be close
}

bench_fit("0")     # pure PyTorch CPU
bench_fit("auto")  # Mojo when available
```

Expect the Mojo run to finish faster (more so at larger `n_sites` /
`n_species` / `sampling`) with a final loss within a few $10^{-3}$ relative
of the torch run -- statistical equivalence, not bitwise identity.

# Troubleshooting

| Symptom | Fix |
|---|---|
| `is_torch_available()` is FALSE | Run `install_diagnostic()`; most often `RETICULATE_PYTHON` points nowhere or at the wrong environment. Set it before `library(sjSDM)` and restart R. |
| Mojo backend seems ignored | Check `file.exists(here::here("work/mojo-backend/mc_grad_server_bin"))`. If missing, redo Step 3 of installation. Environment variables are read at fit time, but a fresh R session is safest after switching. |
| Fit diverges to a constant loss | Unstandardized predictors. Use `scale(env)`. Affects both backends equally. |
| `torch.randn` dimension errors | Fixed in this repository; make sure workers load the patched code (`options(sjSDM.dev = ...)` in parallel CV). |
| Want GPU (Apple MPS)? | `options(sjSDM.device = "mps")`; note the Mojo backend is CPU-only, so MPS fits fall back to torch. |
| pixi or build step fails | Confirm you are inside `work/mojo-backend/` when running `pixi` commands; `pixi run mojo` must resolve the Mojo standard library through the pixi environment. |

# Where to go next

- Package docs: `?sjSDM`, `?bioticStruct`, `?linear`, `?sjSDMControl`,
  `?sjSDM_cv`.
- Benchmarks and parity checks: `Code/benchmarks/`,
  `Code/bench_realdata.R`, `Code/bench_real_occurrence.R`.
- Mojo kernel sources and validation harness:
  `work/mojo-backend/mojo/` and `Feasibility_study.md`.
- Original sjSDM paper: Pichler & Hartig (2021),
  *A new method for faster and more accurate joint species distribution
  modelling* (Methods in Ecology & Evolution).
