---
title: "sjSDM from R: tutorial and Mojo CPU setup"
author: "sjSDM developers"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,   # heavy model fits; knit fast and let readers run chunks
  comment = "#>"
)
```

This tutorial has two parts:

1.  Install and select either the standard PyTorch backend or the optional Mojo CPU backend.
2.  Work through the main sjSDM workflow: simulation, fitting, species associations, prediction, $R^2$, variation partitioning, cross-validation, spatial models, standard errors, and plotting.

Code chunks are not run while this document is rendered. Run them interactively, in order. The basic fit is quick; cross-validation, ANOVA refits, and standard errors take longer.

# Background: what is sjSDM?

A *species distribution model* (SDM) relates species occurrence to the environment. A *joint* SDM (jSDM) models all species at once and estimates **species associations** -- residual correlations between species after accounting for the environment. These associations can indicate biotic interactions, shared unmeasured habitat variables, or dispersal history.

sjSDM implements scalable jSDMs using deep-learning machinery (PyTorch):

- Multivariate probit or logit likelihoods with a **low-rank covariance** for species associations (`bioticStruct(df = ...)`).
- Environmental (and optionally spatial) predictor layers via `linear()` formulas, plus optional deep neural network layers via `DNN()`.
- Fast minibatch-gradient-descent fitting that scales to tens of thousands of sites.

The optional **Mojo backend** replaces the inner Monte Carlo likelihood computation (the hot loop) with a hand-written kernel compiled to native Apple Silicon code. Predictor layers, optimizers, prediction, and Hessian-based standard errors remain in PyTorch.

# Installation

## Who this tutorial is for

This tutorial covers the **Mojo-accelerated CPU build** of sjSDM, which currently requires **Apple Silicon (macOS, arm64)** and a full clone of the development repository at `dougwyu/s-jSDM`.

> **Not on Apple Silicon?** You don't need this tutorial. Install the original, pure-PyTorch sjSDM instead -- it is fully supported on macOS (Intel), Windows, and Linux:
>
> ```{r fallback, eval=FALSE}
> install.packages("remotes")
> remotes::install_github("TheoreticalEcology/s-jSDM/sjSDM")
> library(sjSDM)
> install_sjSDM(version = "cpu")
> install_diagnostic()   # use this detailed report if something is off
> ```
>
> **On Apple Silicon but don't want Mojo?** Stay here anyway. Installing from the upstream repository is currently unreliable on Apple Silicon, whereas this fork contains the compatibility fixes (released as [v0.1.0](https://github.com/dougwyu/s-jSDM/releases/tag/v0.1.0)). Plain PyTorch-mode sjSDM works without any of the pixi/Mojo setup below:
>
> ```{r fallback-apple, eval=FALSE}
> install.packages("remotes")
> remotes::install_github("dougwyu/s-jSDM/sjSDM", ref = "v0.1.0")
> library(sjSDM)
> install_sjSDM(version = "cpu")
> install_diagnostic()   # use this detailed report if something is off
> ```
>
> Everything from Section 4 onward of this tutorial works identically with either install; just skip Sections 2--3 (Mojo setup and switching).

## One install path: clone + pixi

**If everything is already installed, jump to Step 4.**

The Mojo backend lives alongside a project-local Python environment defined with [pixi](https://pixi.sh), a single-binary package manager. PyTorch 2.5.1, Mojo 1.0.0, and MAX 26.5.0 are the versions validated for v0.2.0. The ranges in `pixi.toml` are not an immutable lock, so commit a tested `pixi.lock` before relying on byte-for-byte reproducible installs.

**Step 1 -- clone the repository:**

```{bash clone, eval=FALSE}
git clone https://github.com/dougwyu/s-jSDM.git
cd s-jSDM
```

**Step 2 -- install pixi** (one-time; see <https://pixi.sh> for alternatives):

```{bash pixi, eval=FALSE}
# macOS/Linux:
curl -fsSL https://pixi.sh/install.sh | sh
```

**Step 3 -- create the Python environment, build the Mojo worker, and create the reticulate wrapper.** From the repository root:

```{bash build, eval=FALSE}
cd work/mojo-backend
pixi install
pixi run mojo build -O 3 mojo/mc_logit_grad_server.mojo -o mc_grad_server_bin
.pixi/envs/default/bin/python -m venv reticulate-venv --system-site-packages
```

The `-O 3` flag matters: unoptimized Mojo builds are roughly 50x slower. The thin `reticulate-venv` wrapper is also required: recent reticulate versions do not treat a pixi prefix without a `conda` executable as a usable Conda environment. When this step finishes, both of these paths should exist:

```{bash verify-build, eval=FALSE}
test -x mc_grad_server_bin
test -x reticulate-venv/bin/python
```

**Jump here if everything is already installed**

**Step 4 -- install the R-side development tools, select a backend, and load the package.** Run the package installation once:

```{r r-prerequisites, eval=FALSE}
install.packages(c("devtools", "here"))
```

Then start each R session from the repository and run:

```{r reticulate}
Sys.setenv(RETICULATE_PYTHON = here::here(
  "work/mojo-backend/reticulate-venv/bin/python"
))
Sys.setenv(SJSDM_MOJO_BACKEND = "auto")
stopifnot(file.exists(Sys.getenv("RETICULATE_PYTHON")))

# load_all() runs the REPO code -- including the Python backend in
# sjSDM/inst/python AND its ability to find the Mojo worker binary:
devtools::load_all(here::here("sjSDM"), quiet = TRUE)

is_torch_available()
reticulate::py_config() # use this diagnostic with the pixi setup
```

> **Why not `library(sjSDM)`?** If you install this fork as a regular package, the installed copy looks for the Mojo worker *inside your R library*, won't find it, and silently falls back to PyTorch -- no error, just no acceleration. Just use `devtools::load_all()` on the checkout as shown above.

> **Key idea for R users:** R talks to Python through `reticulate`, which sjSDM handles internally. `RETICULATE_PYTHON` selects the Python environment. The Mojo binary is a separate native worker that sjSDM finds automatically; you do not call it directly.

# Using the Mojo CPU backend

## What it is

The Mojo backend accelerates the Monte Carlo logit/probit likelihood and its analytic gradients on **CPU**. It consists of the two pieces you built during installation:

| Piece | File | Role |
|----|----|----|
| Persistent worker | `work/mojo-backend/mc_grad_server_bin` | Long-lived process computing loss + gradients per training batch |
| R-side bridge | `sjSDM/inst/python/sjSDM_py/mojo_bridge.py` | Feeds tensors/seeds to the worker through a pipe |

You do **not** call either directly. Backend selection is read from Python's environment when a model is built.

## Turning it on / off

Set the backend **before** `devtools::load_all()` starts Python:

```{r mojo-switch}
Sys.setenv(SJSDM_MOJO_BACKEND = "auto") # default: Mojo when eligible
Sys.setenv(SJSDM_MOJO_BACKEND = "1")    # force Mojo; error if unavailable
Sys.setenv(SJSDM_MOJO_BACKEND = "0")    # force PyTorch
```

Once reticulate has started Python, a later `Sys.setenv()` call is not copied into Python's `os.environ`. Restart R after changing it, or deliberately update the active Python process:

```{r mojo-switch-live}
set_sjsdm_backend <- function(backend = c("auto", "mojo", "torch")) {
  backend <- match.arg(backend)
  value <- c(auto = "auto", mojo = "1", torch = "0")[[backend]]
  reticulate::py_run_string(sprintf(
    "import os; os.environ['SJSDM_MOJO_BACKEND'] = '%s'", value
  ))
  invisible(value)
}

set_sjsdm_backend("auto")
```

Auto mode selects Mojo only for a CPU logit/probit fit when the worker binary exists. Batches containing missing responses use PyTorch. Use the default `dtype = "float32"`; v0.2.0 does not automatically fall back for `float64` and instead raises an error.

> **Important v0.2.0 limit:** server-side noise generation initializes at most 262,144 values per request. Until that is fixed, require `sampling * step_size * biotic_rank <= 262144`. For a larger request, use PyTorch or run `Sys.setenv(SJSDM_MOJO_SEED_TRANSPORT = "0")` before `devtools::load_all()`; the latter sends an explicit, fully initialized noise tensor and can be slower.

## When does Mojo help?

The valid end-to-end benchmarks in this repo (`Code/dev/bench_realdata.R`) use `step_size = 10`, so they stay below the v0.2.0 seed-transport limit. They report:

- About 1.9x end-to-end at 10,000 sites x 200 species.
- About 1.3x at 3,000 sites x 60 species.
- Standard errors (`getSe`) always use PyTorch regardless of the setting, because Hessians need second-order autodiff.

The published shape-grid maxima include requests beyond the seed generator's safe limit and should be rerun after that defect is fixed.

Two caveats worth knowing:

1.  Cross-backend results are **statistically equivalent, not bit-identical**: float32 arithmetic differs slightly between the Mojo kernel and PyTorch's einsum, so loss histories agree only to \~$10^{-3}$ relative.
2.  Keep predictors **standardized**. sjSDM fits can diverge on raw-scale covariates (e.g., temperatures near 6000); this affects both backends. `scale(env)` fixes it.

# Data: simulate a community

We create a 200-site, 10-species community with known environmental responses and known species associations, so we can check that sjSDM recovers them.

```{r simulate-data}
set.seed(42)
n_sites   <- 200
n_species <- 10

X <- matrix(rnorm(n_sites * 3), n_sites, 3)
colnames(X) <- c("temp", "precip", "elev")

beta0 <- rnorm(n_species)                               # intercepts
beta  <- matrix(rnorm(3 * n_species), 3, n_species)     # env responses
Sigma <- matrix(0.5, n_species, n_species)
diag(Sigma) <- 1                                       # associations

library(mvtnorm)
logit_p <- cbind(1, X) %*% rbind(beta0, beta)
Z <- rmvnorm(n_sites, sigma = Sigma)
Y <- matrix(rbinom(n_sites * n_species, 1, plogis(logit_p + Z)),
            n_sites, n_species)
```

Alternatively, use sjSDM's own simulator. The predictor matrix is `env_weights`, the response matrix is `response`, and the truth includes `species_weights` and `correlation`:

```{r simulate-SDM, eval=FALSE}
sim <- simulate_SDM(sites = 500, species = 20, correlation = TRUE)
str(sim)
```

Or analyze one of the bundled empirical datasets (e.g. `butterflies`, `eucalypts` -- presence/absence matrices shipped with the package).

# Fitting a basic model

The core function is `sjSDM()`. The three model components map directly to arguments:

```{r fit-basic}
fit <- sjSDM(
  Y = Y,                                        # sites x species 0/1 matrix
  env = linear(data = X, formula = ~ temp + precip + elev),
  biotic = bioticStruct(df = 5),                # rank-5 covariance factor
  family = binomial("logit"),
  iter = 100,                                   # training epochs
  sampling = 100,                               # MC samples per batch
  verbose = FALSE
)

summary(fit)
```

Notes on the knobs:

- `iter`, `sampling`, and `step_size` are the main speed/accuracy trade-offs. If omitted, `step_size` is 10% of the sites. More `sampling` reduces Monte Carlo noise in the likelihood.
- Select a link with `family = binomial("logit")` or `family = binomial("probit")`; Mojo handles both.
- Optimizer, learning-rate scheduling, and early stopping live in `sjSDMControl()` -- see `?sjSDMControl`.
- For non-linear responses, swap `linear(...)` for `DNN(data = X, ...)`.
- Regularize the association structure with, for example, `bioticStruct(df = 5, lambda = 1e-4, on_diag = TRUE)`; tune `lambda` by cross-validation.

# Inspecting species associations

```{r associations}
cov_mat <- getCov(fit)      # estimated species covariance matrix
round(cov_mat, 2)

getCor(fit)                 # covariance rescaled to correlations

# Do we recover the strong positive association we simulated?
mean(cov_mat[upper.tri(cov_mat)])
```

Species weights (the actual fitted parameter matrices) are available via `getWeights(fit)` / `setWeights()` if you need raw access.

## Do fits recover known association strength?

A useful sanity check: simulate two communities that differ **only** in the strength of their true species associations (weak vs. strong off-diagonal covariance), fit the same environmental model to each, and verify that sjSDM's estimated associations track the truth.

```{r assoc-recovery}
simulate_community <- function(assoc_strength, seed = 7) {
  set.seed(seed)  # common random numbers isolate the covariance change
  Sig <- matrix(assoc_strength, n_species, n_species)
  diag(Sig) <- 1
  logit_p_c <- cbind(1, X) %*% rbind(beta0, beta)
  Zc <- mvtnorm::rmvnorm(n_sites, sigma = Sig)
  matrix(rbinom(n_sites * n_species, 1, plogis(logit_p_c + Zc)),
         n_sites, n_species)
}

fit_community <- function(Y_comm, seed = 1) {
  m <- sjSDM(
    Y = Y_comm,
    env = linear(data = X),
    biotic = bioticStruct(df = 5),
    family = binomial("logit"),
    iter = 100, sampling = 100, seed = seed, verbose = FALSE
  )
  list(model = m, partition = summary(anova(m))[, "R2 Nagelkerke"])
}

Y_weak   <- simulate_community(0.1, seed = 7)
Y_strong <- simulate_community(0.6, seed = 7)

res_weak   <- fit_community(Y_weak)
res_strong <- fit_community(Y_strong)

# Variation partitioning should assign more deviance to associations
# when the true associations are strong:
rbind(
  weak_associations   = res_weak$partition,
  strong_associations = res_strong$partition
)

# ... and the mean estimated off-diagonal covariance should be larger:
mean_offdiag <- function(m) {
  C <- getCov(m)
  mean(C[upper.tri(C)])
}
round(c(weak   = mean_offdiag(res_weak$model),
        strong = mean_offdiag(res_strong$model)), 3)
```

You should see the strong-association community show a substantially larger association component in the partition and a larger mean fitted covariance. This is a good template for testing whether a dataset actually supports inference about biotic structure before interpreting it.

## Do fits detect dispersal limitation?

Dispersal limitation leaves a characteristic fingerprint: species occur (or fail to) in **spatially coherent patches** that no measured environmental variable explains. We simulate this by drawing each community's latent site effects from a *spatially autocorrelated* Gaussian field over the coordinates -- weak or strong -- while keeping environmental responses and associations identical. A spatial sjSDM should assign more deviance to the spatial component when dispersal limitation is strong.

```{r dispersal}
# Exponentially decaying spatial covariance over coordinates:
spatial_field <- function(coords, range = 0.5) {
  D <- as.matrix(dist(coords))
  mvtnorm::rmvnorm(1, sigma = exp(-D / range))   # one site-level draw
}

simulate_dispersal_community <- function(dispersal, coords,
                                         n_sp, seed = 1) {
  set.seed(seed)
  # Latent patchiness shared by all species, scaled by dispersal strength:
  field <- drop(spatial_field(coords))
  Z <- dispersal * outer(field, rep(1, n_sp))
  logit_p <- cbind(1, X) %*% rbind(beta0,
                                   matrix(rnorm(3 * n_sp), 3, n_sp))
  matrix(rbinom(nrow(X) * n_sp, 1, plogis(logit_p + Z)),
         nrow(X), n_sp)
}

coords <- matrix(runif(n_sites * 2), n_sites, 2)
SP     <- generateSpatialEV(coords)

fit_spatial <- function(Y_comm) {
  m <- sjSDM(
    Y = Y_comm,
    env = linear(data = X),
    spatial = linear(data = SP, formula = ~ 0 + .),
    family = binomial("logit"),
    iter = 100, seed = 1, verbose = FALSE
  )
  summary(anova(m))
}

part_weak   <- fit_spatial(simulate_dispersal_community(0.2, coords, n_species))
part_strong <- fit_spatial(simulate_dispersal_community(1.5, coords, n_species))

rbind(
  weak_dispersal_limitation   = part_weak[, "R2 Nagelkerke"],
  strong_dispersal_limitation = part_strong[, "R2 Nagelkerke"]
)
```

The spatial component of the partition should grow markedly under strong dispersal limitation, while the pure-environment component stays similar -- the signal is spatial structure *beyond* what the environment explains.

## Do fits detect environmental filtering?

Here, stronger environmental filtering means that occurrence responds more strongly to the measured environmental gradients. We hold everything else fixed and scale the true environmental coefficients. The stronger-filtering community should assign more explained deviance to the environmental component.

```{r filtering}
simulate_filtered_community <- function(beta_scale, seed = 1) {
  set.seed(seed)
  Sig <- matrix(0.1, n_species, n_species); diag(Sig) <- 1
  logit_p_f <- cbind(1, X) %*% rbind(beta0,
                                     beta_scale * beta)
  Zf <- mvtnorm::rmvnorm(n_sites, sigma = Sig)
  matrix(rbinom(n_sites * n_species, 1, plogis(logit_p_f + Zf)),
         n_sites, n_species)
}

fit_plain <- function(Y_comm) {
  m <- sjSDM(
    Y = Y_comm,
    env = linear(data = X),
    biotic = bioticStruct(df = 5),
    family = binomial("logit"),
    iter = 100, sampling = 100, seed = 1, verbose = FALSE
  )
  summary(anova(m))
}

part_weak_filter   <- fit_plain(simulate_filtered_community(0.25))
part_strong_filter <- fit_plain(simulate_filtered_community(2.0))

rbind(
  weak_environmental_filtering   = part_weak_filter[, "R2 Nagelkerke"],
  strong_environmental_filtering = part_strong_filter[, "R2 Nagelkerke"]
)
```

Under strong filtering you should see both a larger total $R^2$ and a partition dominated by the environment-only component. When filtering is weak, the residual association component can absorb more of the unexplained structure. Association estimates can represent omitted environmental or spatial structure as well as genuine biotic interactions.

# Prediction and model fit

```{r predict}
pred <- predict(fit)        # sites x species occurrence probabilities
head(pred[, 1:4])

# Predict at new environmental values:
newdata <- data.frame(temp = c(-1, 0, 1), precip = 0, elev = 0)
predict(fit, newdata = newdata)

Rsquared(fit)               # Nagelkerke-style R^2 for the joint model
```

# Variation partitioning (ANOVA)

`anova()` refits reduced models (environment-only, associations-only) and decomposes explained deviance. This is slower than fitting the full model because of the refits.

```{r anova}
av <- anova(fit)
av
summary(av)                 # includes Nagelkerke R^2 per component
plot(av)                    # barplot of the partition
```

For a spatial model, `internalStructure()` further splits the association component into local and spatially structured parts, and `plotAssemblyEffects()` visualizes species responses along gradients.

# Cross-validation

Use `sjSDM_cv()` to tune regularization strength. Two important details:

- It takes **raw data**, not a fitted model.
- The tuning argument is `tune_steps`, not `IT`.

```{r cv}
options(sjSDM.dev = here::here("sjSDM"))

cv <- sjSDM_cv(
  Y = Y, env = linear(data = X),
  biotic = bioticStruct(df = 5),
  family = binomial("logit"),
  tune = "random", tune_steps = 5,
  CV = 3, iter = 100, sampling = 100,
  n_cores = 2
)

plot(cv)                            # CV curves over lambda
```

Parallel workers load sjSDM themselves; when developing against this repo, point workers at the source checkout with `options(sjSDM.dev = here::here("sjSDM"))`.

# Spatial models

Add spatial predictors (Moran's eigenvector maps from site coordinates):

```{r spatial}
coords <- matrix(runif(n_sites * 2), n_sites, 2)
SP <- generateSpatialEV(coords)          # spatial eigenvectors

fit_sp <- sjSDM(
  Y = Y,
  env = linear(data = X),
  spatial = linear(data = SP, formula = ~ 0 + .),
  family = binomial("logit"),
  iter = 100, seed = 1, verbose = FALSE
)

anova(fit_sp)   # partitions deviance: environment | space | associations
```

# Standard errors and plots

`plot(fit)` draws coefficient estimates with error bars. Computing the error bars requires per-species Hessians, which is the most expensive post-fitting step -- request SEs once with `se = TRUE`, or let `plot()` trigger `getSe()`.

```{r se-plot}
fit_se <- sjSDM(Y = Y, env = linear(data = X),
                biotic = bioticStruct(df = 5),
                family = binomial("logit"),
                se = FALSE, iter = 100, seed = 1, verbose = FALSE)

# getSe() returns an updated model; the matrix is stored in $se.
fit_se <- getSe(fit_se)
se <- fit_se$se

p_coef <- plot(fit_se)      # ggplot2 coefficient plot
p_coef
```

Assembly-effect diagnostics for a spatial ANOVA:

```{r assembly}
intstr <- internalStructure(anova(fit_sp))
plotAssemblyEffects(intstr)
```

# Benchmarking torch vs. Mojo yourself

A quick timing harness compares backends on the same data and model seed. Because Python is already running, it uses the `set_sjsdm_backend()` helper defined earlier rather than `Sys.setenv()`:

```{r bench}
bench_fit <- function(backend) {
  set_sjsdm_backend(backend)
  t0 <- Sys.time()
  m <- sjSDM(Y = Y, env = linear(data = X),
             biotic = bioticStruct(df = 5),
             family = binomial("logit"),
             iter = 100, sampling = 100, seed = 123, verbose = FALSE)
  c(seconds = as.numeric(difftime(Sys.time(), t0, units = "secs")),
    final_loss = tail(m$history, 1))
}

bench_fit("torch")
bench_fit("auto")
set_sjsdm_backend("auto")
```

Run each backend more than once for a serious benchmark and report the distribution of timings. Losses should be close, not bit-identical, because the backends use different random-number generation and float32 reduction order.

# Troubleshooting

| Symptom | Fix |
|----|----|
| `is_torch_available()` is FALSE | Run `reticulate::py_config()`. Most often `RETICULATE_PYTHON` points nowhere or was set after Python started. Set it before loading sjSDM and restart R. |
| Mojo backend seems ignored | Check `file.exists(here::here("work/mojo-backend/mc_grad_server_bin"))`. If missing, redo Step 3. If Python is already active, switch with `set_sjsdm_backend()` or restart R. |
| Large Mojo fit gives suspect results | In v0.2.0, check whether `sampling * step_size * biotic_rank` exceeds 262,144. Use PyTorch or set Python's `SJSDM_MOJO_SEED_TRANSPORT=0` until the generator is fixed. |
| `dtype = "float64"` fails in auto mode | v0.2.0 does not apply the documented auto fallback for float64. Use `dtype = "float32"` with Mojo or force PyTorch. |
| Fit diverges to a constant loss | Unstandardized predictors. Use `scale(env)`. Affects both backends equally. |
| `torch.randn` dimension errors | Fixed in this repository; make sure workers load the patched code (`options(sjSDM.dev = ...)` in parallel CV). |
| Want GPU (Apple MPS)? | `options(sjSDM.device = "mps")`; note the Mojo backend is CPU-only, so MPS fits fall back to torch. |
| pixi or build step fails | Confirm you are inside `work/mojo-backend/` when running `pixi` commands; `pixi run mojo` must resolve the Mojo standard library through the pixi environment. |

# Where to go next

- Package docs: `?sjSDM`, `?bioticStruct`, `?linear`, `?sjSDMControl`, `?sjSDM_cv`.
- Benchmarks and parity checks: `Code/dev/benchmarks/`, `Code/dev/bench_realdata.R`, `Code/dev/bench_real_occurrence.R`.
- Mojo kernel sources and validation harness: `work/mojo-backend/mojo/` and `Feasibility_study.md`.
- Upgrading Mojo/MAX versions: see the "Maintaining the Mojo/MAX pin" section of [Code/README.md](README.md).
- Original sjSDM paper: Pichler & Hartig (2021), *A new method for faster and more accurate joint species distribution modelling* (Methods in Ecology & Evolution).
